---
title: "Mystery Data 1"
author: "Max Israelit"
date: "2/26/2018"
output: 
  word_document:
    reference_docx: manuscriptstyle.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r background, echo=FALSE, message=FALSE, warning=FALSE}
#install.packages(c("magicaxis","scales","wesanderson","plotrix"), repos = "http://cran.us.r-project.org")
library(magicaxis)
library(scales)
library(wesanderson)
library(plotrix)

trees <- read.csv("tree_allometry.csv")
trees2 <- trees[trees$growingCondition == "FW",]
trees2 <- droplevels(trees2)
```

##Investigating Patterns and Drivers of Leaf Strategies  
  
Using the mystery data, I first narrowed down the available dataset to include only those plants that were considered wild and measured in the field. I decided to use specific leaf area (SLA) as a metric of interest in this investigation. Specific leaf area is calculated as the area of a leaf divided by the mass of the leaf. Because individual leaf data was not available, I used the whole-leaf data provided and divided total plant leaf area by total plant leaf mass. Specific leaf area can be interpreted as a plant's investment into each leaf per unit area. Increases in mass while area remains constant usually denotes that the plant is inputting more resources into structure or photosynthetic machinery while maintaining the same area, which is generally a more conservative approach to resource use. This usually means the leaf becomes more expensive and often more effective. The reasons for doing this can be complex, and we will try to explore a few environmental drivers of SLA here.  

##Patterns of SLA by Latitude

Latitude corresponds to a number of environmental conditions that likely influence SLA. For one, latitude is a key determinate of photosynthetically active radiation (PAR), which often correlates to higher possible productivities. This elevated PAR is also available for more of the year near the equator, as seasonal variation is minimal. This, in turn, can lead to longer growing seasons and decrease some of the climactic pressures on plants that exist at higher latitudes.

```{r latitude, echo=FALSE, message=FALSE, warning=FALSE}

trees2$sla <- trees2$a.lf/trees2$m.lf
latcols <- c("chocolate1","chartreuse3","green4","cadetblue2","darkorchid2","orangered2","royalblue1","yellow3","burlywood")
latcols2 <- alpha(latcols, 0.25)
par(mar=c(4,4,1,1), cex.axis=.8, cex.lab=1.1, mgp=c(2.5, 1, 0))
plot(sla ~ latitude, data=trees2, pch=19, cex=1.5, col=latcols2[trees2$vegetation],
     ylab="Specific Leaf Area", xlab="Latitude (Degrees)",
     ylim=c(0,250), xlim=c(-60, 70))
legend("topleft", levels(trees2$vegetation), pch=19, col=latcols, bty="n", cex=1.2)

```
**Figure 1.**Specific Leaf Area by latitude. Legend entries are coded as follows: BorF=Boreal Forest, Gr=Grassland, Sav=Savanna, Sh=Shrubland, TempF=Temperate Forest, TempRF=Temperate Rainforest, TropRF=Tropical Rainforest, TropSF=Tropical Seasonal Forest, Wo=Woodland. Each point represents SLA calculated as an average across the whole plant. Negative latitude values indicate sites south of the equator, while positive values indicate north of the equator.  
  
  
The distribution of SLA across latitude appears bimodal (Figure 1). One peak, slightly north of 20 degrees south, is driven by tropical seasonal forests. The second peak, centered around 45 degrees north, is driven by Temperate Forests and reaches higher than the first peak but is narrower. The relatively high SLAs exhibited in temperate rainforests and tropical rainforests contribute to the broadness of the first peak. This dataset clearly suffers from gaps in the data in unsampled latitudes, but these gaps don't negate the general trends seen here. It appears that SLA remains at a medium to low level in tropical rainforests near the equator. This pattern could be due to limiting factors such as the high plant turnover, limited nutrient availability, or intense spatial competition that characterize these regions. It's also possible that high light availability and favorable growing conditions render investments in leaf area impractical. These drivers would be less notable in tropical seasonal forests and temperate forests, where the temporary nature of leaves in these seasonal forests inhibit great investment in structure. In order to disentangle these influences, I will investigate the impacts of particular environmental conditions on SLA.  
  
##Mean Annual Precipitation
  
Precipitation is a major determinate of water availability for each plant. Water availability has a variety of physiological implications for plants, often occuring at the leaf level. Water loss can occur through stomata or evaporation from leaf tissues, and is necessary to maintain leaf temperatures. It may be possible for greater investments in leaf structure to improve the heat or drought tolerance of photosynthetic tissues.  

```{r map, echo=FALSE, message=FALSE, warning=FALSE}
pftcols <- alpha(c("darkolivegreen", "cadetblue2", "indianred2", "mediumorchid2"), 0.5)
trees.map <- trees2[complete.cases(trees2$map),]
par(mar=c(4,4,1,1), cex.axis=.8, cex.lab=1.1, mgp=c(2.5, 1, 0))
plot(sla ~ map, data=trees.map, pch=19, ylab="Specific Leaf Area", xlab="Mean Annual Precipitation (mm)", 
     cex=1.25, col=pftcols[trees.map$pft],
     ylim=c(0,250), xlim=c(0, 5000))
legend("topright", levels(trees.map$pft), pch=19, col=pftcols, bty="n", cex=1.2, inset=0.1)
```
**Figure 2.** The effects of mean annual precipitation on specific leaf area. Each point represents SLA calculated as an average across the whole plant. Legend entries are coded as follows: DA=Deciduous angiosperm, DG=Deciduous gymnosperm, EA=Evergreen angiosperm, EG=Evergreen gymnosperm.  
  
  
The highest SLA was observed at slightly less than 1000 mm of precipitation annually (Figure 2). However, there was great variation in SLA across the range of MAP values available, which likely reflects differences in niche, canopy localization, and other factors not accounted for in this dataset. in general, areas with higher MAP had lower and less variable SLAs. As MAP increases, more plants make greater investments in structure as opposed to direct areal expansion, with the exception of areas with very low MAP which exhibited low SLA. This trend may be explained by data availability, as perhaps sites were not sufficiently sampled across equivalent conditions for all MAP values. However, in interpreting the trend presented here it appears that greater drought stresses, such as those elicited by lower precipitation levels, lead plants to invest more in their leaf structure relative to areal expansion. This additional investment may come in the form of a thicker cuticle to prevent water loss or other, similar changes. The decreasing trend of SLA with increasing MAP past 2000 mm could be a result of declining productivity, as evidenced by papers discussed in class. In this scenario, lower SLAs are representative of plants seeking to conserve resources under more stressful environmental conditions. Moreover, when looking at plant functional type in particular, deciduous angiosperms seem most susceptible to this trend whereas evergreen angiosperms are more resilient. There was almost no variation in evergreen gymnosperms, likely because these plants are present in boreal forests and environmental conditions are too universal to elicit any variation in leaf strategy.  
  
  
##Mean Annual Temperature
  
Temperature is another environmental and climactic factor that is a major determinate of plant physiology. Plants in warmer areas tend to be more productive and respire more, while plants in colder areas can have far lower metabolisms. On a leaf level, temperature directly influences water stress. Warmer air can force leaves to expend more energy on evaporative cooling. Warmer air also has a higher moisture capacity, which can increase the amount of water lost when stomata are opened. Thus, temperature may have a variety of effects on leaf strategy.  
  
```{r echo=FALSE, message=FALSE, warning=FALSE}
trees.mat <- trees2[complete.cases(trees2$mat),]
par(mar=c(4,4,1,1), cex.axis=.8, cex.lab=1.1, mgp=c(2.5, 1, 0))
plot(sla ~ mat, data=trees.mat, pch=19, ylab="Specific Leaf Area", xlab="Mean Annual Temperature (Celsius)", 
     cex=1.25, col=pftcols[trees.mat$pft],
     ylim=c(0,250), xlim=c(0, 30))
legend("topright", levels(trees.mat$pft), pch=19, col=pftcols, bty="n", cex=1.2, inset=0.1)
```
**Figure 3.** The effects of mean annual temperature on SLA. Each point represents SLA calculated as an average across the whole plant. Legend entries are coded as follows: DA=Deciduous angiosperm, DG=Deciduous gymnosperm, EA=Evergreen angiosperm, EG=Evergreen gymnosperm.  
  
  
As mean annual temperature increases, specific leaf area increases and becomes more variable, with the exception of a tall peak around 5 degrees Celsius (Figure 3). This dataset also suffers from gaps, but in investigating the observed trend, it eappears that plants in hotter areas tend to invest less into leaf structure relative to area. Reasons for this could include that plants at lower temperatures may invest more into structures that would increase leaf survivability or that the limited producitivity in lower temperatures could cause plants to be more conservative with their resources. The alternative may be that colder climates tend to be drier, which would increase rates of water loss. Having smaller leaves may limit rates of water loss. The large peak around 5 degrees Celsius doesn't follow this trend, which is likely because of the limitations of the MAT metric. As MAT is taken as an average across a full year, the low temperature could actually be a simplification of an area with high summer temperatures and low winter temperatures. This peak may then support the observed trend of increasing SLA with increasing MAT, as the leaves would have a high SLA due to high summer temperatures and then would be shed by the deciduous plants for the winter. Similar patterns amongst plant functional types are seen as with MAP, with evergreen gymnosperms existing in a small climactic range and exhibiting little variation, deciduous angiosperms reacting more sensitively to MAT, and evergreen angiosperms reacting more slowly to changes in MAT.  
  
##Plant Functional Type
  
There may be inherent biases towards certain levels of SLA based on a plants' functional type. We have seen notable differences in the sensitivities of plant functional types to climactic changes, so it is important to consider the possibility of any basal biases. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
pftcolssmall <- alpha(c("darkolivegreen", "indianred2", "mediumorchid2"), 0.5)
treespft <- trees2[trees2$pft != "DG",]
treespft <- droplevels(treespft)
par(mar=c(4,4,1,1), cex.axis=.8, cex.lab=1.1, mgp=c(2.5, 1, 0))
boxplot(sla ~ pft, data=treespft, col=pftcolssmall, ylim=c(0,100),
        ylab="Specific Leaf Area", xlab="Plant Functional Type")
```
**Figure 4.** Specific leaf areas of three plant functional types. Each point represents SLA calculated as an average across the whole plant. Legend entries are coded as follows: DA=Deciduous angiosperm, EA=Evergreen angiosperm, EG=Evergreen gymnosperm.  
  
  
The specific leaf areas of each plant functional type are variable (Figure 3). Deciduous angiosperms have the highest basal SLA on average, as well as the greatest variation. Evergreen angiosperms are on average lower in SLA than deciduous angiosperms, and have lower variability. Evergreen gymnosperms have the lowest mean SLA and variability. The low basal SLA in evergreen gymnosperms is likely due to their use of needles instead of leaves as well as the minimal variation in climactic conditions across their main range (boreal forests). Deciduous angiosperms may have a higher SLA on average because their strategy of shedding leaves each year would make investments into a particular leaf less valuable. Though there does appear to be a basal pattern of SLA, the climactic observations made earlier are still visible within each plant functional type and are therefore still valid.  
  
##Leaf Nitrogen
  
Changes in SLA may reflect investment into different leaf structures as well as responses to certain climactic variables. In order to explore how SLA aligns with increases in photosynthetic machinery, I compared levels of SLA to the amount of leaf nitrogen found in each leaf. Leaf nitrogen was presented as total leaf nitrogen across the whole plant, so leaf nitrogen per gram of leaf tissue was calculated on the whole plant scale. It was than log transformed and presented here. 

```{r nitrogen}
trees3 <- trees2[complete.cases(trees2$sla),]
trees3 <- trees3[complete.cases(trees3$n.lf),]
trees3$nitro <- log10((1000*trees3$n.lf)/trees3$m.lf)
nreg <- lm(nitro ~ sla, data=trees3)
r2 <- round(summary(nreg)$r.squared, 2)
r2_var <- expression(R^2)
r2lab <- paste(r2_var, r2, sep = " = ")
par(mar=c(4,4,1,1), cex.axis=.8, cex.lab=1.1, mgp=c(2.5, 1, 0))
plot(nitro ~ sla, data=trees3, pch=19, col=pftcols[trees3$pft],
     ylab="Log Leaf Nitrogen (g per g Tissue)", xlab="Specific Leaf Area",
     ylim=c(-1,7), xlim=c(0,55))
ablineclip(nreg, x1 = min(trees3$sla), x2 = max(trees3$sla), lty=1, lwd=2)
legend("topleft", levels(trees2$pft), pch=19, col=pftcols, bty="n")
text(50, 5.5, paste(expression(R^2), r2, sep= "=")  )
```
**Figure 5.** The relationship between specific leaf area and leaf nitrogen. Both SLA and leaf nitrogen were calculated on a whole-plant basis, using total leaf area, total leaf mass, and total mass of nitrogen in leaves. Legend entries are coded as follows: DA=Deciduous angiosperm, DG=Deciduous gymnosperm, EA=Evergreen angiosperm, EG=Evergreen gymnosperm.  
  
  
This figure illustrates a generally positive correlation between SLA and the log of grams of leaf nitrogen per gram of tissue (Figure 5). However, this relationship is loose and likely insignificant (R-squared = 0.18). SLA is likely not very explanatory for the variation in leaf nitrogen due to several reasons. One is the constrictions of each plant functional type. It appears that each plant functional type is governed by a different relationship between SLA and leaf nitrogen, and were the regression calculated separately for each plant functional type they may have been more powerful. However, such analyses are limited in applicability for the reasons described below, and their incorporation is outside the scope of this study. A second reason for the lack of a significant relationship between SLA and leaf nitrogen is that SLA does only not represent a greater investment into photosynthetic machinery, but can also be a result of climactic pressures as indicated earlier in this report. Thus, a relationship between leaf nitrogen and SLA may be visible if all other environmental conditions are kept constant. Nonetheless, this supports the concept that SLA is at least partially determined by climactic and environmental factors, and represents differences in plant strategy and physiology based on various state factors.



