---
title: "Mystery Data Set 1"
author: PJ Murphy
Date: Feburary 27, 2018
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The following are five graphs depicting analysis of a mystery data set 1, which includes tree allometry information from across the globe. First, the relationship of both leaf mass fraction and stem mass fraction versus total plant mass is examined, with vegetation type included as a third variable. Next, the differences between nitrogen content and mass in three types of wood is explored in samples from Temperate Forests.

```{r plot 1, echo=FALSE}
trees <- read.csv("tree_allometry.csv")
library(magicaxis)
trees$leafmf<-trees$m.lf/trees$m.to
trees2 <- droplevels(trees[complete.cases(trees$leafmf),])
trees3 <- droplevels(trees2[complete.cases(trees2$leafmf) & 
                             complete.cases(trees2$m.to) & 
                            complete.cases(trees2$vegetation),])
vegnames<-c("Boreal Forest", "Grassland", "Shrubland", "Temperate Forest",
 "Temperate Rainforest", "Tropical Rainforest",
 "Tropical Seasonal Forest", "Woodland")
log_m.to<- log10(trees3$m.to)
plot(leafmf~log_m.to,data=trees3, axes=FALSE, xlab="Total Plant Mass (kg)",
     ylab="Leaf Mass Fraction (kg/kg)", pch=1, col=c("red", "orange", "lightcoral",
  "forestgreen", "purple", "gold", "black", "blue")[trees3$vegetation])
magaxis(side=1, unlog=1, frame.plot=TRUE)
axis(2)
legend("topright",legend=vegnames, pch=1,col=c("red", "orange", "lightcoral",
"forestgreen","purple", "gold","black", "blue"), cex=0.6)
```
*Figure 1.* Relationship of leaf mass fraction (kg/kg) to total plant mass (kg) from various global vegetation types. Some vegetation types were excluded from the analysis based on their lack of plant mass data. 

It is apparent that as total plant mass size increases, leaf mass fraction decreases, which is consistent with the findings of current literature (Poorter et al. 2015). The data from the tropical rainforest (gold) predominates the figure, but it is noted that the total plant size of these rainforest plants is localized, whereas the information concerning temperate forests (green) and boreal forests (red) is much more variable. This is interesting because tropical rainforests are known for being rich in biodiversity, and one would think there would be a wider range of sizes of these plants. While, there is not much variability in terms of total plant size for tropical rainforest plants, some of the highest and lowest values regarding plant mass fraction are seen here. This is most likely related to the fact that some plants are under the canopy in the tropics and must rely on sunflecks, as light can be a limiting factor in their growth. To make the most of these sunflecks, they need massive leaves to absorb as much light energy as possible when given the chance. 


```{r plot 2, echo=FALSE}
trees <- read.csv("tree_allometry.csv")
trees$stemmf<-trees$m.st/trees$m.to
trees2 <- droplevels(trees[complete.cases(trees$stemmf),])
trees3 <- droplevels(trees2[complete.cases(trees2$stemmf) & 
                              complete.cases(trees2$m.to) & 
                              complete.cases(trees2$vegetation),])
vegnames<-c("Boreal Forest", "Grassland", "Shrubland", "Temperate Forest",
            "Temperate Rainforest", "Tropical Rainforest",
            "Tropical Seasonal Forest", "Woodland")
log_m.to<- log10(trees3$m.to)
plot(stemmf~log_m.to,data=trees3, axes=FALSE, xlab="Total Plant Mass (kg)",
     ylab="Stem Mass Fraction (kg/kg)", pch=1, col=c("red", "orange", "lightcoral",
                                                     "forestgreen", "purple", "gold", "black", "blue")[trees3$vegetation])
magaxis(side=1, unlog=1, frame.plot=TRUE)
axis(2)
legend("topleft",legend=vegnames, pch=1,col=c("red", "orange", "lightcoral",
                                               "forestgreen","purple", "gold","black", "blue"), cex=0.4)
```
*Figure 2.* Stem mass fraction (kg/kg) compared with total plant mass (kg), in relation to vegetation type where the samples are found. Again, vegetation types which lacked total plant mass data were excluded from the figure. 

Figure 2 exemplifies the concept that as trees grow bigger, they partition more to stem growth. It is seen that there is an increase in stem mass fraction as total plant mass increases. This makes sense intuitively since as trees gain more mass, they are usually getting taller, meaning there is more stem mass. As in Figure 1, the tropical rainforest samples (gold) are localized in terms of total plant mass, whereas temperate forest (green) samples are more varaible. It can be concluded from this that there is more varaibility in the temperate forests than in tropical rainforest in terms of total plant size and stem mass fraction.    

```{r plot 3, echo=FALSE}
trees <- read.csv("tree_allometry.csv")
trees5 <- droplevels(trees[complete.cases(trees$m.ss),])
trees6 <- droplevels(trees5[complete.cases(trees5$m.ss) &
            complete.cases(trees5$n.ss),])
logm.ss<- log10(trees6$m.ss)
logn.ss<- log10(trees6$n.ss)
plot(logn.ss~logm.ss, data=trees6, axes=FALSE, xlab= "Mass of Sapwood (kg)",
     ylab= "Sapwood Nitrogen Content (kg/kg)", pch=0, col=c("saddlebrown"))
magaxis(side=1, unlog = 1, frame.plot=TRUE)
axis(2)

```
*Figure 3.* Mass of sapwood including branches (kg) compared to sapwood nitrogen content (kg/kg) for plant samples found in temperate forests. 

Although, there is some variability in the data, it can be seen in Figure 3 that as mass increases so does nitrogen content. As mass of sapwood, and therefore total tree mass increase, it requires more resources like nitrogen to manage and support larger individuals. 
```{r plot 4, echo=FALSE}
trees <- read.csv("tree_allometry.csv")
trees7 <- droplevels(trees[complete.cases(trees$m.sh),])
trees8 <- droplevels(trees7[complete.cases(trees7$m.sh) &
        complete.cases(trees7$n.sh),])
logm.sh<- log10(trees8$m.sh)
logn.sh<- log10(trees8$n.sh)
plot(logn.sh~logm.sh, data=trees8, axes=FALSE, xlab= "Mass of Heartwood (kg)",
     ylab= "Heartwood Nitrogen Content (kg/kg)", pch=0, col=c("saddlebrown"))
magaxis(side=1, unlog = 1, frame.plot=TRUE)
axis(2)
```
*Figure 4.* Relatioship between mass of heartwood including branches (kg) compared to heartwood nitrogen content (kg/kg) for tree samples from temperate forests.

The interaction of nitrogen content and mass in the heartwood of these plants shows that the bigger the tree, the more nitrogen it has partitioned to it's stem. Heartwood is the center of the trees stem composed of many dead cells, and has plays a mainly structural role. So more the bigger the tree, the more heartwood it will need to structurally support itself, and therefore must partition more resources, including nitrogen, to heartwood.  

```{r plot 5, echo=FALSE}
trees <- read.csv("tree_allometry.csv")
trees9 <- droplevels(trees[complete.cases(trees$m.sb),])
trees10 <- droplevels(trees9[complete.cases(trees9$m.sb) &
    complete.cases(trees9$n.sb),])
logm.sb<- log10(trees10$m.sb)
logn.sb<- log10(trees10$n.sb)
plot(logn.sb~logm.sb, data=trees10,axes=FALSE, xlab= "Mass of Bark (kg)",
     ylab= "Bark Nitrogen Content (kg/kg)", pch=0, 
     col=c("saddlebrown"))
magaxis(side=1, unlog = 1, frame.plot=TRUE)
axis(2)
```
*Figure 5.* Mass of bark including branches compared to nitrogen content of bark for trees found in temperate forests. 

Interestingly, bark was found to have the highest levels of nitrogen for any type of stem wood. Bark is responsible for defending the live cells of the sapwood from insects, animals, and disease, and thus it is necessary to devote more resources here, nitrogen being one of these resources. 
