---
output:
  word_document: default
  html_document: default
---
u---
author: Shana Shapiro
date: 2/13/2018
output: 
  word_document:
    reference_docx: manuscriptstyle.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Allometry 

In this mystery dataset, I looked at the allometric relationships from the provided tree database. There are some interesting relationships between variables like total mass, leaf mass fraction, stem mass fraction, and more. 

#Plot 1
Broad relationship between area of stem at breast height and the aboveground total mass 
```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=6, fig.height=6}
trees <- read.csv("tree_allometry.csv")
library(plotrix)
library(magicaxis)
library(scales)
library(wesanderson)
model <- lm(log10(m.so) ~ log10(a.stbh), data = trees)
mycol <- alpha("aquamarine4", 0.5)

plot(log10(m.so) ~ log10(a.stbh), data = trees, ylab = "Aboveground Total Mass (kg)", 
     xlab = "Area of Stem at Breast Height (m2)", col = mycol, pch = 16, axes=FALSE)
magaxis(side=c(1,2), unlog=c(1,2), frame.plot=TRUE)
```
This graph shows the broad relationship between the area of a plant stem at breast height and the aboveground total mass. The area of the stem at breast height indicates the size of a tree. Since larger cross sections require more support and more biomass to hold up the tree, the aboveground total mass will also increase. 


#Plot 2 
Stem Mass versus Leaf Mass with respect to Growing Condition 
```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=6, fig.height=6}
mycols <- (c("thistle4", "hotpink4", "green4", "deepskyblue4", "aquamarine3", "orchid4", "yellowgreen"))
mycols2 <- alpha(mycols, 0.5)

plot(log10(m.st) ~ log10(m.lf), data = trees, ylab = "Leaf Mass (kg)", xlab = "Stem Mass (kg)", 
     col = mycols2, pch = 16, axes = FALSE)
magaxis(side=c(1,2), unlog = c(1,2), frame.plot = TRUE)
legend("bottomright", levels(trees$growingCondition), pch = 16, col = mycols2, bty = "o", inset = 0.06)
```
This graph shows the stem mass as a function of the leaf mass, differentiated by the conditions in which the trees grow. In general, as the stemm mass increases, the leaf mass also increases. As the stems increase in size, the tree is larger and the tree grows more leaves. As a result, larger trees also have a larger total of leaf mass. 


#Plot 3 
Stem Mass and Leaf Mass with respect to Plant Functional Type 
```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=6, fig.height=6}
thecols <- c("aquamarine", "blue", "green", "red")
thecols2 <- alpha(mycols, .75)

plot(log10(m.lf) ~ log10(m.st), data=trees, xlab="Stem Mass (kg)", ylab="Leaf Mass (kg)",
     axes=FALSE, pch=16,col=thecols2)
magaxis(side=c(1,2), unlog=c(1,2), frame.plot=TRUE)
legend("bottomright", legend=levels(trees$pft), pch=16, col=thecols2, bty='o',inset=0.02)
```
Similar to the previous plot, this relationship shows leaf mass as a function of stem mass, but with respect to plant functional types. In this graph, we see that the functional types are present throughout the curve, suggesting that the relationship between stem mass and leaf mass is consistent across all functional types. 


#Plot 4 
Leaf Mass Fraction and Total Mass with respect to Plant Functional Type 
```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=6, fig.height=6}
trees$lmf <- (trees$m.lf / trees$m.to)

plot(lmf ~ log10(m.to), data=trees, xlab="Total Mass (kg)", ylab="Leaf Mass Fraction (kg kg-1)",
     xaxt="n", pch=16,col=mycols2)
magaxis(side=c(1), unlog=c(1), frame.plot=TRUE)
abline(h=0.3544533)
legend("bottomright", legend=levels(trees$pft), pch=16, col=mycols2, bty = 'o',inset=0.02)
```
This graph shows the leaf mass fraction as a function of the total mass by plant functional type. The relationship corrects for total mass, so it specifically focuses on the allocation of leaf mass in relation to the rest of the plant. On this graph, the mean leaf mass fraction of 0.354 is plotted. As seen on the plot, the data is distributed fairly equally across the line for smaller total mass, suggesting the leaf mass allocation is variable for different plants of various funcitonal types. 



#Plot 5
Stem Mass Fraction and Total Mass with respect to Plant Functional Type 
```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=6, fig.height=6}
trees$smf <- (trees$m.st / trees$m.to)

plot(smf ~ log10(m.to), data=trees, xlab="Total Mass (kg)", ylab="Stem Mass Fraction (kg kg-1)", xaxt="n", pch=16,col=mycols2)
magaxis(side=c(1), unlog=c(1), frame.plot=TRUE)
abline(h=0.3343945)
legend("bottomright", legend=levels(trees$pft), pch=16, col=mycols2, bty = 'o',inset=0.02)
```
This graph shows the relationship between the stem mass fraction as a function of the total mass with respect to plant functional type. The mean stem mass fraction of 0.334 is plotted. As seen on the graph, the data clusters together at a smaller stem mass fraction for smaller plants and also at a larger stem mass fraction for larger trees. This suggests a consistent stem mass fraction depending on the size of the plant. 


