---
title: "Mystery Data # 2 "
author: "Cheyenne Brown"
date: "March 5, 2018"
output: 
  word_document:
    reference_docx: manuscriptstyle.docx
---
``` {r Figure 1, echo=FALSE, warning=FALSE, message=FALSE}
leaves <-read.csv("leafarea_global.2csv")
leafdata <- leaves[complete.cases(leaves$Leaf.size_cm2),]
library(magicaxis)
par(mar=c(5,5,1,1))
syms <- c(15,16,17)
mycols <- c("powderblue", "lightcoral", "seagreen")
plot(Leaf.size_cm2~Elevation..m., data = leaves, col=mycols[TaxonGroup], xlab= "Elevation of Plant (m)", ylab = "Leaf Size (cm2)", xmin=c(0, 5265), ylim=c(0,2562), pch=syms[Decid_or_Ever])

label3 <- levels(leaves$Decid_or_Ever)
leglab <- c(label3, levels(leafdata$TaxonGroup))
legend("topright", legend = leglab, pch=c(syms,15,15,15), inset = 0.01, col = c("black", "black", "black", mycols), bty='n')
```
\
Figure 1. Leaf Size (cm2) Based on Elevation of Plant (m)
\
In Figure 1, leaf size is measured in centimeters squared (cm2) and elevation is measured in meters. Based on the legend, the black symbols are used to show whether the plant is a deciduous type (the "D" represented by the black circle), evergreen (the "E" represented by the black triangle), or neither represented by the black square. Therefore, Deciduous, Evergreen, or neither is simply represented by the symbol type, whereas taxon group is represented by color. The angiosperms (angio on the legend) is powder blue, the ferns are light coral in color, and the gymnosperms (gymno on the legend) is sea green in color. 
\
As seen in Figure 1, the majority of plants in this data set are angiosperms, with minimal ferns and gymnosperms. Additionally, there is a cluster of plants at lower elevations and many plants at low elevations have low leaf sizes, with the exception of some outliers. Evidently, there are certainly deciduous species, evergreen species, and neither deciduous nor evergreen species present in this data set. Most of the ferns are found at elevations less than 1000 meters and have a leaf size of roughly 500 cm2. The gymnosperms have leaf sizes between 0cm2 and 1cm2 but can be seen at elevations between 1000 meters and 2000 meters, roughly midway between 2000 meters and 3000 meters, and roughly midway between 3000 meters and 4000. with this being said, there are very few gymnosperms in this sample. The majority of angiosperms are found at elevations below 3000 meters and the majority have a leaf size of 1000 cm2 or less (also with certain outliers). There are very few plants at high elevations to begin with and few with large leaf sizes. 

```{r Figure 2, echo=FALSE, warning=FALSE, message=FALSE}
leaves <-read.csv("leafarea_global.2csv")
leafdata <- leaves[complete.cases(leaves$Leaf.size_cm2),]
cols <- c("cornflowerblue", "darkorchid1", "deeppink")
library(maps)
map('world')
par(mar=c(1,1,3,3))
points(leaves$Latitude~leaves$Longitude, pch=16, col=cols[leafdata$TaxonGroup], cex=sqrt(leaves$Leaf.size_cm2/1000))
leg <- levels(leafdata$TaxonGroup)
legend("bottomleft", pch=16, pt.bg = cols, col=c(cols), legend = leg, xpd=TRUE, cex=1.3, inset=-.1, bty='n')
```
\
Figure 2. How Leaf Size (cm2) Varies Around the World Based on Taxon Group. 
\
In Figure 2, Leaf Size is measured in centimeters squared (cm2) and Latitude is measured in decimal degrees.Figure 2 takes longitude and latitude coordinates to plot a map of angiosperms, ferns, and gymnosperms across the globe. The size of leaves of each taxon group corresponds to the size of the points plotted on the map. The taxon groups in Figure 2 include angiosperms (labeled as "angio" and represented by cornflower blue circles in the legend of Figure 2), ferns (represented by dark orchid circles in the legend and map), and gymnosperms (labeled as "gymno" and represented by deep pink circles in the legend and map). 
\
\
\
\
\
\
\
\
\
\
\
\
```{r Figure 3, echo=FALSE}
par(mar=c(5,5,1,1))
plot(Leaf.size_cm2~Latitude, data=leaves, xlim=c(-55,76), ylim=c(0,2562), xlab="Latitude (decimal degrees)", ylab="Leaf Size (cm2)", col=cols[TaxonGroup], pch=16)
legend("topleft",legend = leg, pch=16, inset=0.01, col = cols, bty = 'n')
```
\
Figure 3.
Leaf Size of Three Taxon Groups Based on Latitude
\
In Figure 3, Leaf Size is measured in centimeters squared (cm2) and Latitude is measured in decimal degrees. The taxon groups in Figure 3 include angiosperms (labeled as "angio" and represented by cornflower blue circles in the legend of Figure 3), ferns (represented by dark orchid circles in the legend), and gymnosperms (labeled as "gymno" and represented by deep pink circles in the legend). 
\ 
\
\
```{r Figure 4, echo=FALSE}
par(mar=c(5,5,1,1))
plot(Leaf.size_cm2~MAP, data=leaves, xlim=c(0,6580), ylim=c(0,2562), xlab="Mean Annual Precipitation (mm)", ylab="Leaf Size (cm2)", col=cols[TaxonGroup], pch=16)
legend("topright",legend = leg, pch=16, inset=0.01, col = cols, bty = 'n', cex=1.1)
```
\
Figure 4.
Leaf Size of Three Taxon Groups Based on Mean Annual Precipitation
\
In Figure 4, Leaf Size is measured in centimeters squared (cm2) and Mean Annual Precipitation is measured in millimeters (mm). The taxon groups in Figure 4 include angiosperms (labeled as "angio" and represented by cornflower blue circles in the legend of Figure 4), ferns (represented by dark orchid circles in the legend), and gymnosperms (labeled as "gymno" and represented by deep pink circles in the legend). 
\
\
\
        Figure 2, Figure 3, and Figure 4 are all interconnected. Figure 2 is a depiction of what is graphed in Figure 3, simply in the form of a map. Figure 2 exhibits Leaf Size based on geographic location (Longitude and Latitude). On the other hand, Figure 3 plots what can be seen in Figure 2, but rather in a numerical format in terms of Leaf Size. Figure 3 similarly shows how Leaf Size changes based on geographic location, specifically Latitude. Finally, Figure 4 has a similar trend to that of Figure 3, which demonstrates how Mean Annual Precipitation is correlated with Latitude and impacts the Leaf Size of plants across the world. With regards to Figure 2, it is evident that there are more angiosperms found in this data set, with a few gymnosperms (one visibly in the United States, one in South America, and a few in or close to Australia). The angiosperms are particularly large in Central America and the northern part of South America. However, it appears that there are angiosperms found on each continent within this data set, but many are found in Asia, South America, Central America, Africa, and a few in the United States.With regards to the legend, angio stands for angiosperms and gymno corresponds to gymnosperms.Similar to Figure 2, there are more angiosperms seen in both Figure 3 and Figure 4 than the two other Taxon Groups. I would imagine that mean annual precipitation varies based on location, hence why latitude is linked to certain amounts of precipitation and the leaf size of plants. The idea that location impacts leaf size also corresponds and relates to Figure 1—precipitation varies upon elevation and elevation varies upon latitude, and leaf size varies upon these three factors (precipitation, elevation, and latitude). When looking at latitude in Figure 3, it is important to consider that 0 decimal degrees marks the equator and in Figure 3, it can be seen that latitudes close to the equator have larger leaf sizes. Figure 2 can create a visualization of the relationship between latitude and leaf size because it is evident that locations near the equator have larger leaf sizes, such as countries in Central America and in the Northern regions of South America. Furthermore, one would expect there to be greater precipitation close to the equator because there is cycling of water as temperatures and humidity are higher. Leaf size can be dependent upon the amount of water available and the distribution of this liquid. Therefore, in Figure 4, there is a cluster of plants with leaf sizes below 500 cm2 and mean annual precipitation ranging between roughly 100 to 4000. Within this data set, high mean annual temperatures are not as common and in the event of high mean annual precipitation (above 6000mm in Figure 4 for example), the leaf size is small. It seems that large leaf sizes can occur when there is not too much precipitation yet not too little precipitation. Therefore, I would imagine having a balance with regards to mean annual precipitation would create a conducive environment for leaves to grow. 