---
title: "An analysis of leaf area"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Overall there was no general trend between the leaf areas and the geographic location of the sites they were found in (Figure 1). Transpiration in plants is the process by which water is transferred from the surface of a plant leaf into the air. Equilibrium transpiration implies that this is the rate where condensation onto the leaf is occurring at the same rate as water is being transpired off. Equilibrium transpiration did not appear to change with increasing leaf size (Figure 2). Leaf size increased with increasing mean annual precipitation until a certain point where it decreased (Figure 3). This may imply that after a certain point the plant becomes over saturated and is no longer increasing leaf size with increasing precipitation. The trend between leaf size and mean annual temperature was highly dependent on growth form (Figure 4). In some cases it showed a positive correlation and in others it was only positively correlated until a certain point when it began to decrease. This may reflect how different carbon allocation strategies creating the different growth forms react to temperature.

```{r maps, echo=FALSE, warning=FALSE, message=FALSE}
leaves<-read.csv("leaf_area/leafarea_global.csv")

library(maps)
library(rworldmap)
library(RColorBrewer)
  map('world')
library(classInt) 
mapcolors <- brewer.pal(6, "Blues")
brks<-classIntervals(leaves$Leaf.size_cm2, style="fixed",
                     fixedBreaks=c(0, 0.5, 1, 1.5, 2.0, 2.6))
bins<- brks$brks
leaves$ls_bins <- cut(leaves$Leaf.size_cm2, breaks = bins)
points(leaves$Longitude, leaves$Latitude, pch=21, bg=mapcolors[leaves$ls_bins]) 
leaflabel <- expression(Leaf~size~~(cm^2))
legend("topleft", pch=21, pt.bg=mapcolors, legend=bins, title=leaflabel, 
       xpd=TRUE, inset=-.1)
```

Figure 1. Map depicting leaf areas (cm^2^) across study sites. 



```{r evap, echo=FALSE, warning=FALSE}
library(scales)
library(wesanderson)
vegfac <- length(levels(leaves$Growth_form))
vegcols <- wes_palette('Zissou', vegfac, type = 'continuous')
vegcols2 <- alpha(vegcols, .25)
leaflabel <- expression(Log~leaf~size~~(cm^2))
par(mar = c(4, 7, 2, 7))
plot(log10(Leaf.size_cm2)~ETq, data=leaves, xlab= "Equillibrium evapotranspiration (mm)", ylab=leaflabel, pch=16, col= vegcols[Growth_form])
legend("topright", legend=levels(leaves$Growth_form), pch=16, col=vegcols, bty='n', inset=c(-.7,.02), cex=0.8, xpd=TRUE)



```

Figure 2. The relationship between individual logged leaf size and equilibrium transpiration (mm). The colors represent the growth form of that plant. 

```{r growthMAP, echo=FALSE, warning=FALSE}
vegfac <- length(levels(leaves$Growth_form))
vegcols <- wes_palette('Zissou', vegfac, type = 'continuous')
vegcols2 <- alpha(vegcols, .25)
leaflabel <- expression(Log~leaf~size~~(cm^2))
par(mar = c(4, 7, 2, 7))
plot(log10(Leaf.size_cm2)~MAP, data=leaves, xlab= "Mean annual precipitation (mm)", ylab=leaflabel, pch=16, col= vegcols[Growth_form])
legend("topright", legend=levels(leaves$Growth_form), pch=16, col=vegcols, bty='n', inset=c(-.7,.02), cex=0.8, xpd=TRUE)

```

Figure 3. The relationship between individual logged leaf size and mean annual precipitation (mm) at that site. The colors correspond to the growth form of that plant. 
```{r growthMAT, echo=FALSE, warning=FALSE}
vegfac <- length(levels(leaves$Growth_form))
vegcols <- wes_palette('Zissou', vegfac, type = 'continuous')
vegcols2 <- alpha(vegcols, .25)
leaflabel <- expression(Log~leaf~size~~(cm^2))
par(mar = c(4, 7, 2, 7))
plot(log10(Leaf.size_cm2)~MAT, data=leaves, xlab= "Mean annual temperature (C)", ylab=leaflabel, pch=16, col= vegcols[Growth_form])
legend("topright", legend=levels(leaves$Growth_form), pch=16, col=vegcols, bty='n', inset=c(-.7,.02), cex=0.8, xpd=TRUE)

```

Figure 4. The relationship between individual logged leaf size and mean annual temperature at that site. The colors correspond to the growth form of that plant. 


