---
title: "Leaf Area Analysis"
author: "Karl Brown"
date: "3/4/2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
leaf<-read.csv("leafarea_global.csv")
leaf2<-leaf[complete.cases(leaf$Whole_leaf_size_cm2),]
library(plotrix)
library(maps)
library(RColorBrewer)
library(classInt)
library(scales)
```

```{r,echo=FALSE}
map(xlim=range(leaf$Longitude), ylim=range(leaf$Latitude))
brks<-classIntervals(leaf2$Whole_leaf_size_cm2,style="fixed",fixedBreaks=c(0,2000,4000,6000,8000)) 
bins<- brks$brks
colors <- brewer.pal(5, "Greens")
leaf2$leaf_bin <- cut(leaf2$Whole_leaf_size_cm2, breaks = bins)
points(leaf$Longitude, leaf$Latitude, pch=21, bg=colors[leaf2$leaf_bin],cex=sqrt(leaf2$Whole_leaf_size_cm2/800))
legend("bottomleft", pch=21, pt.bg=colors, legend=bins, title="Whole Leaf Area (cm^2)", 
       xpd=TRUE, inset=-.1,cex=.6)
```

**Figure 1** Global map of the distribution of whole leaf area size. Both circle size and circle color indicate the size of the leaf. 

It is clear that leaf area is generally small with a few exceptions in Australia and Africa where leaf size is exceptionally large. It can also be noted that the areas where leaf size is larger tends to be around the equator. The connection between leaf size and distance from the equator will be explored further later.

```{r,echo=FALSE}
colly<-c("chartreuse","cornflowerblue","orangered")
collys=alpha(colly, .5)
lereg<-lm(leaf$Leaf.size_cm2~leaf$Elevation..m.)
plot(leaf$Leaf.size_cm2~leaf$Elevation..m.,xlab="Elevation (m)",ylab="Leaf Size (cm^2)",col=collys[leaf$Decid_or_Ever],pch=16)
legend("topright",legend=c("Annual","Deciduous","Evergreen"), pch=16,col=collys,cex=1.1)
ablineclip(lereg,x1=min(min(leaf$Elevation..m.),x2=max(leaf$Elevation..m.)),lwd=2)
```

**Figure 2** Negative relationship between Leaf size and elevation with distinction of plant type (annual, deciduous, or evergreen).

While the sample size of leaves at high elevation are limited, a general negative trend between these two variables can be observed. This relationship may be due to the fact that places of low elevation could be valleys where there is a higher quality and quantity of soil whereas the higher elevations could be areas where topography limits the success of trees and large leaves. 

```{r,echo=FALSE}
lmapreg<-lm(leaf$Leaf.size_cm2~leaf$MAP)
plot(leaf$Leaf.size_cm2~leaf$MAP,col=collys[leaf$Decid_or_Ever],pch=16,xlab="Mean Annual Precipitation (mm)",ylab="Leaf Size (cm^2)")
legend("topright",legend=c("Annual","Deciduous","Evergreen"), pch=16,col=collys,cex=.8)
ablineclip(lmapreg,x1=min(leaf$MAP,na.rm = TRUE),x2=max(leaf$MAP,na.rm=TRUE),lwd=2)
```

**Figure 3** No strong relationship between Leaf Size and Mean annual precipitation with distinction of plant type (annual, deciduous, or evergreen). 

The trend line indicates an increase in leaf size with increasing MAP, but a closer look at the graph reveals that there is no obvious or strong positive relationship. Instead, this figure is interesting in that it shows that there is no real change in leaf size with increasing MAP. There is a great deal of debate surrounding how MAP effects NPP, and while this graph does not deal with NPP, leave size is likely connected to NPP as trees with higher NPP would presumably have larger leaves. Thus this graph and the lack of a relationship between MAP and leaf size further supports the idea that NPP is not influenced by MAP. 

```{r,echo=FALSE}
plot(leaf$Leaf.size_cm2~leaf$Latitude,xlab="Latitude",ylab="Leaf Size (cm^2)",col=collys[leaf$Decid_or_Ever],pch=16)
mean(leaf$Leaf.size_cm2,na.rm=TRUE)
abline(h=40.99085,lty=2,lwd=2,col="black")
legend("topleft",legend=c("Annual","Deciduous","Evergreen"), pch=16,col=collys)
```

**Figure 4** Leaf size is maximized around the equator, with distinction of plant type (annual, deciduous, or evergreen).. The mean leaf size is marked by the dashed line. 

Aside from a group of deciduous trees around latitude 60, leaf size tends to get smaller as latitude moves away from 0. Most of the points that exceed the mean are centered around the equator. This may be due to a lack of samples from higher and lower latitudes, but this data shows that trees around the equator will have larger leaves. This may be due to the close proximity of the sun at the equator, thus trees develop larger leaves in this area to capture more of the available sunlight. 
