---
title: "Global Distribution and Effects of State Factors on Whole Leaf Area for Simple and Compound-Leaved Species"
author: "Marie Pugliese"
date: "March 6, 2018"
output:
  word_document:
   reference_docx: manuscriptstyle.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


```{r, echo=FALSE}
leaves <- read.csv("leafarea_global.csv")
leaves2 <- leaves[,c("Whole_leaf_size_cm2", "Latitude", "Longitude",
                     "Compound_Simple", "Elevation..m.", "MAT")]
leaves3 <- leaves2[complete.cases(leaves2$Whole_leaf_size_cm2) &
                     complete.cases(leaves2$Latitude) &
                     complete.cases(leaves2$Longitude) &
                     complete.cases(leaves2$Compound_Simple) &
                     complete.cases(leaves2$Elevation..m.) &
                     complete.cases(leaves2$MAT), ]

library(maps)
library(scales)

wholearea <- leaves3$Whole_leaf_size_cm2
latitude <- leaves3$Latitude
compoundsimple <- leaves3$Compound_Simple
elevation <- leaves3$Elevation..m.
wholearealab <- expression(Leaf~Area~~(cm^2))
latlab <- expression(Latitude~~(degree))
elelab <- expression(Elevation~~(m))
matlab <- expression(Mean~Annual~Temperature~~(degree~C))
cslab <- c("Compound", "Simple")
cscols <- c("blue", "pink")
cscols2 <- alpha(cscols, .50)
```

#Leaf Area, Distribution, and State Factors

\ When making interpretations based on climatic or distribution-related factors, it is important the data set is all-encompassing. That way, there are no hidden biases. The 11,459 samples used in the data set are from plants found all over the world (Figure 1). Though they do not span every single biome, they do cover many, as well as a broad range of latitudes (Figure 2). It is sufficient.

\ Though plants with leaves of small areas exist at virtually all latitudes, plants with leaves of larger areas generally are found only in the tropics (Figure 2). One explanation as to why this tropics-only pattern is the case is climate, which is largely dictated by precipitation and temperature. Mean annual precipitation (MAP) was plotted against leaf area; however, it did not correlate with leaf area. Much variation in leaf area exists for locations with a low or medium MAP. However, only plants with leaves of small areas were found in places with a high MAP (above 5000 mm). As previous studies have suggested, this decrease of productivity in excess is likely due to decreased nitrogen availability which may occur when water is in excess (Schuur & Matson, 2001).

\ A positive correlation between mean annual temperature (MAT) and leaf area exists. Though plants with leaves of smaller areas are found at a broad range of temperatures, plants with leaves of larger areas are generally found only in places with a high MAT (Figure 3). It is a common pattern in ecology for more activity to occur with increasing temperature. Perhaps in these warmer environments, such as the topics, the increased activity may allow for more growth and thus larger leaf areas.

\ Another state factor, topography (elevation), may also help to explain differences in leaf area sizes. Though plants with leaves of small areas are found at a broad range of elevations, plants with leaves of larger areas are generally found only in places of low elevation (Figure 4). At some point, leaves can no longer be large. Perhaps this negative trend is due to water availability. Runoff works in favor of plants at low elevations, as water will flow down to them but not up and against gravity. With more access to water, a resource, these plants will likely be able to invest more energy into leaf area.


```{r, echo=FALSE}
par(mar=c(0,0,0,0))
map('world')
points(leaves3$Longitude, leaves3$Latitude,
       cex=sqrt(leaves3$Whole_leaf_size_cm2/1000),
       pch=21, bg=cscols2[compoundsimple])
legend("bottom", legend=cslab, pch=16, col=cscols, inset=0.00, cex=0.8)
```

\ **Figure 1.** World distribution map of locations from which sample leaves used in the data set (n=11,459) were taken. Point size corresponds to whole leaf size of the individual. Blue and pink points represent data from compound (n=519) and simple (n=10,940) leaves, respectively.


```{r, echo=FALSE}
par(mar=c(5,5,1,1))
plot(wholearea ~ latitude, data=leaves3,
     ylim=c(0,3500), xlab=latlab, ylab=wholearealab,
     pch=21, col=cscols[compoundsimple])
legend("topright", legend=cslab, pch=21, col=cscols, inset=0.02)
```

\ **Figure 2.** Leaf area (demonstrated by whole leaf size) as a function of latitude. Blue and pink points represent data from compound (n=519) and simple (n=10,940) leaves, respectively. It is important to note that for greater ease of visualizing the data, the y-axis was shortened, resulting in the removal of three data points. Though one exists at a latitude close to 0 degrees, the other two exist at latitudes of about 30 degrees south. However, given that they are two points in a set of 11,500, they are considered outliers.


```{r, echo=FALSE}
par(mar=c(5,5,1,1))
plot(wholearea ~ MAT, data=leaves3,
     ylim=c(0,3500), xlab=matlab, ylab=wholearealab,
     pch=21, col=cscols[compoundsimple])
legend("topleft", legend=cslab, pch=21, col=cscols, inset=0.02)
```

\ **Figure 3.** Leaf area (demonstrated by whole leaf size) as a function of mean annual temperature (MAT). Blue and pink points represent data from compound (n=519) and simple (n=10,940) leaves, respectively. It is important to note that for greater ease of visualizing the data, the y-axis was shortened, resulting in the removal of three data points. All three exist in places with relatively higher MATs.


```{r, echo=FALSE}
par(mar=c(5,5,1,1))
plot(wholearea ~ elevation, data=leaves3,
     ylim=c(0,3500), xlab=elelab, ylab=wholearealab,
     pch=21, col=cscols[compoundsimple])
legend("topright", legend=cslab, pch=21, col=cscols, inset=0.02)
```

\ **Figure 4.** Leaf area (demonstrated by whole leaf size) as a function of elevation. Blue and pink points represent data from compound (n=519) and simple (n=10,940) leaves, respectively. It is important to note that for greater ease of visualizing the data, the y-axis was shortened, resulting in the removal of three data points. All three exist at low elevations.


#Works Cited

Schuur, E. & Matson, A. (2001). Net Primary Productivity and Nutrient Cycling across a Mesic to Wet Precipitation Gradient in Hawaiian Montane Forest. *Oecologia, 128*(3), 431-442.

