---
title: "Assignment 3: Variations in Leaf Area on a Global Scale"
author: "MaryannWebb"
date: "3/1/2018"
output: word_document
---

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
```

## Vegetation type and leaf size 

Comparing leaf area to vegetation type can act as a predictor of what sorts of trees, plants, and environments leaf areas are related to. Figure 1 shows that the mean leaf area of Evergreen and Deciduous trees are roughly the same for our data set. Deciduous and Evergreens are often found in the same areas and are known to survive in similar climate and soil conditions, thus their leaf areas would be similar. Annual plants consist mostly of grasses, which are often found in soil and climate conditions different than tree-d forests. Grasses have different relationships with mycorrhiza than trees and different limiting nutrients, meaning energy allocation is going to be different for annual leaves than evergreen and deciduous trees. 

```{r vegetation, echo=FALSE, warning=FALSE, message=FALSE}
leaves <- read.csv("leaf_area/leafarea_global.csv")
library(magicaxis)
library(scales)
library(plotrix)
ylabel<-expression(Log~of~Leaf~Area~~(cm^2))
levels(leaves$Decid_or_Ever)[1] <- "A"
leaves$log_la <- log10(leaves$Leaf.size_cm2)
par(mar=c(5,5,1,1))
plot(log_la ~ Decid_or_Ever, data=leaves,
     ylab=ylabel, xlab="Vegetation Type", col="darkolivegreen4")

```

Figure 1: Average leaf size (cm^2) by vegetation types. Vegetation types are separated by Annuals (A), Deciduous (D), and Evergreens (E).

## Elevation and Vegetation Type

Climate and elevation are linked to each other as topography can have intense impacts on small areas. Figure 2 shows that there is no obvious link between elevation and leaf area. At low elevations there is a large range of leaf areas. This variation make sense as forests are often more densely population at lower elevations, and the greater the density, the larger the variation in leaf size will be in order to catch as much light as possible in the midst of high competition. 
/
Vegetation type was an additional element in Figure 2 in to see if patterns between elevation and leaf area could be influenced by whether the sample was annual, deciduous, or evergreen. At lower elevations there does not appear to be any sort of pattern between vegetation type, leaf area, and elevation. As elevation increases and samples are more sparsely spread, the samples are mainly annual plants. Annual plants are often cultivators of high altitudes because the condition at these levels can be harsh and are not able to support many large trees.  
```{r elevation, echo=FALSE, message=FALSE, warning=FALSE}
colors<- c("darkolivegreen4", "darkorange4", "darkgray")
colors1<-alpha(colors, .5)
par(mar=c(5, 5, 1, 1))
plot(log10(Leaf.size_cm2)~Elevation..m., data=leaves, 
     xlab="Elevation(m)", ylab=ylabel, yaxt='n', col=colors1[Decid_or_Ever], pch=18)
magaxis(side=2, unlog=2, frame.plot=TRUE)
legend("topright", levels(leaves$Decid_or_Ever), pch=18, col=colors, inset=0.05)

```

Figure 2: Leaf Area (cm^2) by Elevation. Data also visualizes different vegetation types. Vegetation types are separated by Annuals (A), Deciduous (D), and Evergreens (E).



## Precipitation and Vegetation Type

Mean annual precipitation (MAP) is often considered to be a large factor in determining how plants allocate biomass. Figure 3 looks at the relationship between MAP and leaf area to see if there is a relationship between the variables. Like elevation, when MAP is low, there is a wide variation of leaf areas. This is an interesting trend because wetter places, such as tropical rain forests, are considered to have higher diversity than dry areas. Even adding the variable of vegetation type does not shed light on why these dryer areas have a large range of leaf areas. There is a trend as MAP increases, so do the mean leaf areas of a sample. Respiration and photosynthesis are water limited, thus it makes sense that as water becomes less of a limiting factor, plants can afford to increase leaf biomass. The largest gradient of leaf areas is between 4000 and 5000 mm of precipitation annually. This likely represents the tropical biomes where plant diversity is high and leaf size adaptations and differences are necessary in order to capture sunlight. 

```{r precipitation, echo=FALSE}

plot(log_la ~ MAP, data=leaves,
     xlab="Mean Annual Precipitation", ylab=ylabel, pch=18, col=colors1[Decid_or_Ever])
legend("topright", levels(leaves$Decid_or_Ever), pch=18, col=colors, inset=0.05)
```


Figure 3: Average Leaf area compared to mean annual precipitation. Data also visualizes vegetation types within this relationship. Vegetation types are separated by Annuals (A), Deciduous (D), and Evergreens (E).




## Visualizing Leaf Area on a Global Scale

The data set used in this analysis was at a global scale. Figure 4 gives a spatial interpretation of how leaf area differs across the world, and also shows where samples are from. The map shows that in tropical zones, especially south and central America, leaf area tends to be larger than in the poles. Because light is very abundant in these equatorial locations, leaves are larger in order to make the most of the available sunlight. Soils are also not very nutrient rich for extended periods of time in tropical locations as turnover is very fast. The lack of a stable pool of soils in the nutrients means plants will depend more on the energy produced during photosynthesis and expend less energy capturing and storing nutrients from the soil.  
```{r, mapping, echo=FALSE, warning=FALSE, message=FALSE}
library(maps)
library(readr)
par(mar=c(1,1,1,1))
map('world')
points(leaves$Latitude~leaves$Longitude, pch=21, col="darkolivegreen", 
       cex=sqrt(leaves$Leaf.size_cm2/1000))

```

Figure 4: World trends of leaf area (m^2). Point sizes increase where mean leaf area increases. 