---
title: "Factors Affecting Leaf Area"
author: "Josh Winward"
date: "March 8, 2018"
output: 
  word_document:
    reference_docx: manuscriptstyle.docx

---

```{r setup, include=FALSE}
leafarea<-read.csv("leafarea_global.csv")
library(maps)
```


The first visualization of this data set is geographical distribution of all the data points colored by growth type. I chose the relationship between growth type and geographical location because I was interested in where all of these data points came from and how representative of all biomes they are. As the data show, most of data points collected area from trees, ferns, grasses, and shrubs. For these four classes this data set is remarkably complete, samples are collected from all six continents and from areas ranging from coastal to desert and forested areas. It is also interesting to note from this map that by looking at the types of plants sampled on the map one can quite accurately guess the biome which exists there. For example we see that in central Africa there are a high number of grasses and trees in the middle of the continent however towards the coasts it turns to more shrubs indicating that there is likely a forest or grassland in the middle of the continent which turns to more deserted areas as you move towards the coast. This visualization of the data gives a good overview of the geographical scale covered by this data set as well as the difference in nieches occupied by various plant growth types.

```{r echo=FALSE}
types<-levels(leafarea$Growth_form)
colors<-c("darkgreen", "deepskyblue", "darkolivegreen", "forestgreen", "darkseagreen", "darkslategray", "darkorange", "gold", "darkslategray1", "chartreuse", "aquamarine")
map('world')
points(leafarea$Longitude, leafarea$Latitude, pch = 21, bg = colors[leafarea$Growth_form])
legend("left", pch=21, pt.bg=colors, legend=types, title="Growth Form", 
       xpd=TRUE, bty = "n", cex = 0.5, inset = -0.08)
```

Figure 1: A map of the world wherein each data point is plotted based on the geographical location from which it was sampled. Each data point is colored by its plant growth form to demonstrate the variety of plants sampled as well as the different biomes in which each growth form exists. 

##Leaf Area vs Elevation

The second visualization of data I chose to look at was the impact that elevation has on the leaf area. As the data depict the majority of plants can be found in lower elevations and within each elevation there is a huge amount of variability. However, it seem to be that at lower elevations for each plant growth form they are more likely to have a larger leaf surface area. Additionally trees tended to have amongst the highest leaf areas, however the plants with the highest leaf area were the palms which makes sense given their physiology. Most likely the reason for this relationship stems from the fact that plants want to conserve their resources. At lower elevations having a high leaf surface area is beneficial because the tree or plant can more readily photosynthesize, the larger the leaf the more photosynthesis. However the larger the leaf the greater the surface area to volume ratio of said leaf. At lower elevations this does not seem to be a limiting factor, however at higher elevations it seems that plants want to keep their leaves smaller. This is likely because with a higher surface area to volume ratio there is more space available for gas exchange. At higher elevations there are often stronger winds and lower partial pressures of gasses. These aspects of living at higher elevations would incentivize plants to have smaller leaves 1) to protect themselves from higher winds and 2) to prevent excess gas exchange and loss due to the differences in partial pressures of gas at higher altitudes. 

```{r echo = FALSE}
plot(leafarea$Whole_leaf_size_cm2~leafarea$Elevation..m., ylab =  expression("Leaf Area cm" ^"2"), xlab = "Elevation (m)",  pch = 21, bg = colors[leafarea$Growth_form])
legend("topright", pch=21, pt.bg=colors, legend=types, title="Growth Form", 
       xpd=TRUE, inset=.01, bty = "n", cex = 0.65)
```

Figure 2: Leaf area as a function of elevation is depicted. Each data point is colored by the plant growth type as depicted in the figure legend. 

## Leaf area vs Average Daily Irradiance

The third relationship which is explored is the impact that irradiance has on leaf area. We already know that productivity increases as light availability increases, as such I was curious to see if the same pattern exists for Leaf Area. For light availability and productivity, the relationship is positive until the point where it becomes damaging i.e once light reaches a point where it is actually damaging to the plant the productivity goes down. Given the plants want to maximize the amount of resources that they can take in I expected to see that once irradiance reaches a certain saturation point that there would be a maximal leaf size and that any increase in irradiance past that would not incentivize plants to increase their leaf area to take advantage of the increased light as it may become damaging. What I found through plotting these data was that for the vast majority of plant growth types, the amount of light did not have  significant effect on the leaf area. However, for trees and herbs there did not seem to be a clear correlation, rather there were certain irradiance points that caused individuals to increase their leaf area to take advantage of that increase. We see this at 120, 155, and 210. This indicates to me that for a plant to invest in increasing its leaf size there needs to be a certain saturation point of light to make that investment worth it to the plant. If the light irradiance does not get to that point then it is not worth it to the plant to invest in additional leaf area as the increase in light would not lead to enough of an increase in energy production to justify the increased infrastructure.
```{r echo=FALSE}
plot(leafarea$Whole_leaf_size_cm2~leafarea$RADann,  ylab =  expression("Leaf Area cm" ^"2"), xlab = expression("Average Daily Irradiance W/m"^"2"), pch = 21, bg = colors[leafarea$Growth_form])
legend("topright", pch=21, pt.bg=colors, legend=types, title="Growth Form", 
       xpd=TRUE, inset=.01, bty = "n", cex = 0.6)
```

Figure 3: Leaf area as a function of the average daily irradiance. Data points are separated by plant growth type as specified by the figure legend. 

## Leaf Area vs. Relative humidity

The fourth and final relationship depicts leaf area and the average relative humidity in a given area. Before making this plot I thought that as humidity increases the leaf size would decrease so that the amount of water lost to the environment would be minimized. In fact the opposite relationship is shown, as relative humidity increases, the leaf area increases as well at what appears to be an exponential rate. As plant get to around 70% relative humidity the leaf area stays relatively constant. However, past that point the leaf area increases at an exponential rate for trees. After further consideration this relationship appears to make sense. Areas which have an average relative humidity above 80% are typically rain forests. In rain forests there is often staunch competition for resources such as water and light. As such it would make sense that in these locations plants would want to maximize their chance of getting their share of resources. Maximizing leaf area would mean that these plants would be able to get more light, as well as direct more rain to their roots through throughflow. These energetic and nutrient requirements are what I believe has led to the increase in leaf area in areas with high relative humidity. 

```{r echo=FALSE}
plot(leafarea$Whole_leaf_size_cm2~leafarea$RHann,  ylab =  expression("Leaf Area cm" ^"2"), xlab = "Relative Humidity (%)", pch = 21, bg = colors[leafarea$Growth_form])
legend("topleft", pch=21, pt.bg=colors, legend=types, title="Growth Form", 
       xpd=TRUE, inset=.01, bty = "n", cex = 0.65)    
```

Figure 4: Leaf area of plants of different growth forms (as indicated by the figure legend) is depicted for various relative humidities. Relative humidity is measured in percentage and ranges from 30-90% humidity and all plant leaf types (simple and compound) are accounted for. 