---
title: "Analyzing the Effects of Elevated Levels of CO2 in FACE Experiments"
author: "Karl Brown"
date: "4/16/2018"
output:
  word_document:
    reference_docx: "~/Desktop/face/manuscriptstyle.docx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r,echo=FALSE,message=FALSE}
library(doBy)
library(lubridate)
library(sciplot)

wood_biomass<-read.csv("~/Desktop/face/face_woodbiomass.csv")
litter<-read.csv("~/Desktop/face/face_litter.csv")
photo<-read.csv("~/Desktop/face/face_photosynthesis.csv")
photo$Date <- paste(photo$Date,"01", sep="_")
photo$Date <- as.Date(photo$Date, format="%b_%Y_%d")
photo_co2<-summaryBy(Anet~Date+treatment, data=photo,FUN=c(mean,se),keep.names = TRUE)
litter$Date <- as.Date(litter$Date, format="%d/%m/%Y")
litter$year <- as.factor(year(litter$Date))
litter$treatment<-as.factor(ifelse(litter$ring==2|litter$ring==3|litter$ring==6,"aCO2","eCO2"))
line.num <- which.max(litter$twig)
litter <- litter[-line.num,]
line.num <- which.max(litter$twig)
litter <- litter[-line.num,]
ring_diameter <- 25 
ring_area <- pi * (ring_diameter/2)^2
c_frac <- .47
kg_to_g <- 1000
total_wood_ring <- summaryBy(biomass_kg ~ year + Ring + treatment,
                             data=wood_biomass, FUN=sum, na.rm=TRUE, keep.names = T)
total_wood_ring$Date <- paste(total_wood_ring$year, "/12/31", sep="")
total_wood_ring$Date <- as.Date(total_wood_ring$Date, format = "%Y/%m/%d")
dates <- unique(total_wood_ring$Date)
dates <- dates[order(dates)]
wood_npp <- total_wood_ring[total_wood_ring$Date != dates[1],]
wood_npp$Start_date <- wood_npp$Date
for (i in 1:length(wood_npp$Date)) {
  wood_npp$Start_date[i] <- dates[which(dates == wood_npp$Date[i]) - 1] 
  wood_npp$prev_biom_kg[i] <- total_wood_ring$biomass_kg[total_wood_ring$Ring == wood_npp$Ring[i] &
                                                           as.numeric(total_wood_ring$Date-wood_npp$Start_date[i])==0]
}
  wood_npp$wood_growth <- ((wood_npp$biomass_kg - wood_npp$prev_biom_kg)* c_frac * kg_to_g)/ring_area
  wood_npp$timelength <- as.numeric(wood_npp$Date - wood_npp$Start_date)
  wood_npp$wood_yr <- ifelse(wood_npp$timelength == 365, wood_npp$wood_growth/1, wood_npp$wood_growth/2)
  
```

Under higher levels of atmospheric carbon dioxide, it is hypothesized that trees will exhibit higher levels of photosynthesis. This hypothesis is supported by Figure 2 as said figure clearly shows that at all points in time the photosynthesis levels at sites with elevated levels of carbon dioxide are higher than the photosynthesis levels at sites with atmospheric levels of carbon dioxide. Carbon dioxide is a necessary component for the process of photosynthesis, so it makes sense that with more carbon dioxide, there would be more photosynthesis. Elevated photosynthesis levels is a major finding from FACE experiments because in the presence of high levels of carbon dioxide, it is paramount that trees are able to to absorb this carbon dioxide at higher levels. 

```{r, Figure 1, message=FALSE,echo=FALSE}
plot(photo_co2$Anet.mean~photo_co2$Date,type='n',xlab=NA,ylab="Foliage Photosynthesis (umols m^-2 s^-1)")
points(Anet.mean~Date,data = photo_co2[photo_co2$treatment=="eCO2",],type='b',col="blue")
points(Anet.mean~Date,data = photo_co2[photo_co2$treatment=="aCO2",],type='b',col="red")
legend("topright",pch=16,legend=levels(photo_co2$treatment),title="Treatment",col=c("red","blue"))
with(photo_co2, arrows(Date, Anet.mean, Date, Anet.mean+Anet.se, angle=90,
                      length=0.03))
with(photo_co2, arrows(Date, Anet.mean, Date, Anet.mean-Anet.se, angle=90,
                       length=0.03))
```

**Figure 1** Foliage photosynthesis (umols m^-2 s^-1) vs. time with standard errors and distinction between elevated CO2 plots and atmospheric CO2 plots. 
\
\
\
\
\
\

There was never a point that photosynthesis levels of the trees in the elevated carbon dioxide plots were lower than the photosynthesis levels in the atmospheric carbon dioxide plots. This is displayed in Figure 2 as the response ratio, the ratio of photosynthesis in eCO2 plots over photosynthesis in aCO2 plots, never drops below 1. This reinforces that idea presented in Figure 1 that photosynthesis levels are higher in eCO2 plots.

```{r, Figure 2, message=FALSE,echo=FALSE}
eco3<- droplevels(photo_co2[photo_co2$treatment %in% "eCO2",])
colnames(eco3)[3]<-"eanet"
aco3<- droplevels(photo_co2[photo_co2$treatment %in% "aCO2",])
colnames(aco3)[3]<-"aanet"
resp_rat<-merge(eco3,aco3,by="Date")
resp_rat$rr<-resp_rat$eanet/resp_rat$aanet
plot(rr~Date,data=resp_rat,col="blue",type='l',xlab=NA,ylab="Response Ratio (eCO2/aCO2)")
```

**Figure 2** Response Ratio vs. time. Response ratio is defined by photosynthesis levels of eCO2 plots divided by photosynthesis levels of aCO2 plots. 
\
\
\
\
\
\

Total literfall from trees in aCO2 and eCO2 plots was very similar and if anything, literfall from aCO2 plots was higher (Figure 3). Despite changes to photosynthesis and perhaps NPP and biomass, levels of literfall between the groups remain relatively the same, thus literfall will not be significant impacted as atmospheric CO2 continues to rise. 
```{r, message=FALSE, echo=FALSE}
litter$all=litter$twig+litter$bark+litter$seed+litter$leaf
litters<-summaryBy(all~year+treatment, data=litter,FUN=c(mean),keep.names = TRUE,na.rm=TRUE)
litters2 <- droplevels(litters[litters$year !=2020,])
b=barplot(litters2$all,col=c("red","chartreuse"),ylab="Total Literfall (grams)")
barplot(litters2$all,col=c("red","chartreuse"),ylab="Total Literfall (grams)")
mtext(side=1, at=c(1.3,3.7,6.1), text=c("2021", "2022", "2023"), line=3)
axis(1, at=b, labels=litters2[,2])
legend("topleft",pch=16, legend=levels(litter$treatment),col=c("red","chartreuse"))
```

**Figure 3** Total literfall (grams) vs. time with distinction between aCO2 and eCO2 plots. Literfall is composed of twigs, bark, leaves, and seeds. 
\
\
\
\
\
\

Annual net primary production was higher is aCO2 plots than in eCO2 plots in 2021, but these results varied in 2022 and 2023 (Figure 4). While higher NPP in aCO2 plots is surprising, this does not take into account belowgroundgrowth which some researchers believe is where trees in eCO2 expend their chemical energy. Overall, the variability from year to year in NPP displayed in figure 4 reveals that there is still more to learn regarding the effects of elevated CO2 on NPP.

```{r,message=FALSE,echo=FALSE}
courtiscool<-summaryBy(wood_growth~year+treatment, data=wood_npp,FUN=mean,keep.names = TRUE,na.rm=TRUE)
b=barplot(courtiscool$wood_growth,col=c("red","chartreuse"),ylab="ANPP (g C m^-2 yr^-1)")
barplot(courtiscool$wood_growth,col=c("red","chartreuse"),ylab="ANPP (g C m^-2 yr^-1)")
mtext(side=1, at=c(1.3,3.7,6.1), text=c("2021", "2022", "2023"), line=3)
axis(1, at=b, labels=courtiscool[,2])
legend("topright",pch=16, legend=levels(litter$treatment),col=c("red","chartreuse"),title="Treatment")
```

**Figure 4** Annual net primary production(g C m^-2 yr^-1) vs. time with distinction between aCO2 and eCO2.
\
\
\
\
\
\

Literfall remains relatively constant throughout all the months in the experiment except for two occasions in which it was much higher: January 2022 and and October 2023 (Figure 5). It is highly likely that during these two times there was an insect infestation that caused a massive defoliation event that increased litterfall.

```{r,message=FALSE,echo=FALSE}
litter$month=month(litter$Date)
litterz<-summaryBy(leaf~year+month, data=litter,FUN=c(mean),keep.names = TRUE,na.rm=TRUE)
barplot(litterz$leaf,col="chartreuse",ylab="Litterfall (grams)")
mtext(side=1, at=c(0.0,7.1,20,35.1), text=c("Oct. 2020", "Jan. 2021", "Jan2022","Jan. 2023"), line=1,cex=.67)

```

**Figure 5** Total literfall (grams) by month from October 2020 to December 2023.
