---
title: "Mystery Data 4"
author: "Matthew Murphy"
date: "4/20/2018"
output:
 word_document:
   reference_docx: manuscriptstyle.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r plot 1, echo=FALSE}
photo <- read.csv("face_photosynthesis.csv")

photo$Date <- paste(photo$Date, "01", sep="_")
photo$CDate <- as.Date(photo$Date, format= "%b_%Y_%d")

photo_co2 <- summaryBy(Anet ~ CDate + treatment, data=photo, FUN=c(mean, se))

photolab <- expression(A[n]~~(umols~m^-2~s^-1))
treatleg <- c(expression(eCO[2]), expression(aCO[2]))
treatcol <- c("springgreen","palevioletred1")

eANET <- data.frame(photo[,c(2,3,7,8)])
eANET <- droplevels(eANET[eANET$treatment !="aCO2",])
colnames(eANET)[3] <- "eAnet"
aANET <- data.frame(photo[,c(2,3,7,8)])
aANET <- droplevels(aANET[aANET$treatment !="eCO2",])
colnames(aANET)[3] <- "aAnet"

rr <- merge(eANET, aANET, by="CDate")
rr$response_ratio <- rr$eAnet/rr$aAnet

rr_co2 <- summaryBy(response_ratio ~ CDate, data=rr, FUN=c(mean, se))

plot(Anet.mean ~ CDate, data=photo_co2, ylab=photolab, xlab="Date",
     ylim=c(14, 35), type='n', las=2)
points(Anet.mean ~ CDate, data=photo_co2[photo_co2$treatment=="eCO2",], type='b', 
       bg="springgreen", pch=21)
points(Anet.mean ~ CDate, data=photo_co2[photo_co2$treatment=="aCO2",],  type='b', 
       bg="palevioletred1", pch=21)
with(photo_co2, arrows(CDate, Anet.mean, CDate, Anet.mean+Anet.se, angle=90, 
                       length=0.03))
with(photo_co2, arrows(CDate, Anet.mean, CDate, Anet.mean-Anet.se, angle=90, 
                       length=0.03))
legend("topright", legend=treatleg, pch=21, pt.bg=treatcol, inset=0.02, cex=1.2)
```
**Figure 1:** Annual net photosynthesis increases in elevated CO2 environment 


```{r plot 2, echo=FALSE}
photo <- read.csv("face_photosynthesis.csv")

photo$Date <- paste(photo$Date, "01", sep="_")
photo$CDate <- as.Date(photo$Date, format= "%b_%Y_%d")

photo_co2 <- summaryBy(Anet ~ CDate + treatment, data=photo, FUN=c(mean, se))

photolab <- expression(A[n]~~(umols~m^-2~s^-1))
treatleg <- c(expression(eCO[2]), expression(aCO[2]))
treatcol <- c("springgreen","palevioletred1")


eANET <- data.frame(photo[,c(2,3,7,8)])
eANET <- droplevels(eANET[eANET$treatment !="aCO2",])
colnames(eANET)[3] <- "eAnet"
aANET <- data.frame(photo[,c(2,3,7,8)])
aANET <- droplevels(aANET[aANET$treatment !="eCO2",])
colnames(aANET)[3] <- "aAnet"

rr <- merge(eANET, aANET, by="CDate")
rr$response_ratio <- rr$eAnet/rr$aAnet

rr_co2 <- summaryBy(response_ratio ~ CDate, data=rr, FUN=c(mean, se))


plot(response_ratio.mean ~ CDate, data=rr_co2, type='b', 
     ylim=c(1.05,1.55), xlab="Date", ylab="Response Ratio")
with(rr_co2, arrows(CDate, response_ratio.mean, CDate, 
                    response_ratio.mean+response_ratio.se, angle=90, length=0.03))
with(rr_co2, arrows(CDate, response_ratio.mean, CDate, 
                    response_ratio.mean-response_ratio.se, angle=90, length=0.03))
```
**Figure 2:** Response ratio (eCO2/aCO2) predicted to remain positive in future years. 


```{r plot 3, echo=FALSE}
lit_annual <- read.csv("lit_annual.csv")
litter_co2 <- summaryBy(twig + bark+ seed + leaf ~ year + treatment, 
                        data=lit_annual, na.rm=TRUE, keep.names = TRUE)
litter_fum <- droplevels(litter_co2[litter_co2$year !=2020,])
litter_fum$totalflux <- with(litter_fum, twig+bark+seed+leaf)


mycol<-c("palevioletred1", "springgreen")

b<-barplot(litter_fum$totalflux, col=mycol, 
           ylab="Annual Litter Production (g)", ylim=c(0,350))
mtext(side=1, at=c(1.3,3.7,6.1), text=c("2021", "2022", "2023"), line=3)
axis(1, at=b, labels=litter_fum[,2])
box()
```
**Figure 3:** Annual litter production slightly decreased in elevated CO2 environment. 


```{r plot 4, echo=FALSE}
wood <- read.csv("wood_npp.csv")
littera <- read.csv("lit_annual.csv")

littera2 <- droplevels(littera[littera$year !=2020,])
colnames(wood)[2] <- "ring"

anpp <- merge(wood, littera2, by=c("year", "ring","treatment"))
anpp$anpp <- anpp$wood_yr + anpp$twig + anpp$bark + anpp$seed + anpp$leaf

anpplab <- expression(ANPP~~(g~C~m^-2~yr^-1))

boxplot(anpp ~ year*treatment, data=anpp, ylab=anpplab, axes=FALSE, 
        col=c("palevioletred1", "springgreen"))
axis(1, at=c(1.5,3.5,5.5), labels=c("2021", "2022", "2023"))
axis(2)
box()
legend("topright", legend=c("aCO2", "eCO2"), pch=21, pt.bg=c("palevioletred1", "springgreen"), inset=0.02)
```
**Figure 4:** ANPP fluctuates within elevated CO2 environment 


```{r plot 5, echo=FALSE}
litter <- read.csv("face_litter.csv")

basket <- .2 
c_frac <- .47 

litter$CDate <- as.Date(litter$Date, format= "%d/%m/%Y")
litter$year <- as.factor(year(litter$CDate))
litter$treatment <- as.factor(ifelse(litter$ring == 2 |
                                       litter$ring == 3 |
                                       litter$ring == 6,
                                     "aCO2", "eCO2"))

line.num <- which.max(litter$twig)
litter <- litter[-line.num,]
line.num <- which.max(litter$twig)
litter <- litter[-line.num,]
litter_bugs <- summaryBy(leaf ~ CDate + treatment, data=litter, FUN=sum, na.rm=TRUE, 
                         keep.names = TRUE)
polydate <- c(as.Date("2021-12-12"), as.Date("2022-03-13"), as.Date("2022-09-15"), 
              as.Date("2022-11-12"), as.Date("2023-02-11"), as.Date("2023-04-17"), 
              as.Date("2023-09-11"), as.Date("2023-11-23"))
polycol <- alpha("grey", 0.5)

mycol<-c("palevioletred1", "springgreen")

plot(leaf~CDate, data=litter_bugs, type='n', xlab="", ylab="Leaf Litter Biomass (g)")
points(leaf~CDate, data=litter_bugs[litter_bugs$treatment=="eCO2",], type='l', 
       col="palevioletred1")
points(leaf~CDate, data=litter_bugs[litter_bugs$treatment=="aCO2",], type='l', 
       col="springgreen")
legend("topleft", legend=c("aCO2", "eCO2"), lty=1, col=mycol, inset=0.02)
```
**Figure 5:** Defoliation events are exacerbated under elevated CO2 environment. 

##Photosynthesis

Annual net photosynthesis rates are expected to increase within elevated CO2 environments (Fig. 1). With more carbon available for uptake, it is unsurprising that photosynthesis rates would increase. However, while the rate increases, this increase is finite and relatively small, as made evident by the small values for response rations (Fig. 2). These increases likely vary due to availability of other resources, such as water. It would appear these nutrients cycle every two years, which would result in the fluctuations in response ratios we observe. 

##Litterfall

Surprisingly, annual litter production slightly decreases in elevated CO2 environments (Fig. 3). This would imply that the litter components, such as leaves, are not only more productive under these conditions, but they also live longer, reducing the litter produced each year. 

##ANPP

In examining annual net primary production we found there was no clear pattern of influence in the elevated CO2 environment (Fig. 4). During certain years ANPP was higher in elevated CO2 environments, while on other years it was equal or lower in an elevated CO2 environment. This suggests other factors affect ANPP levels more severely than carbon. 

##Problems with Pests

Defoliation events caused by bugs can be a problem at this experiment site. Our findings suggest these events would be exaggerated within elevated CO2 environment (Fig. 5). Leaf litter biomass increased in elevated CO2 environments during each predicted defoliation event. Since photosynthesis rates increase, this is not surprising, as an increase in photosynthesis likely results in an increase in production of biomass. 

