---
title: "FACE"
author: "Alicia Violette"
date: "4/15/2018"
output:
  word_document:
    reference_docx: manuscriptstyle.docx
---

\
By looking at the ways that this forest responds to elevated levels of CO~2~ we are able to get a sense of the ways that forest may respond to the increased levels of CO~2~ that we are projected to experience with climate change. We should also be able to get a sense of the forest's capacity to act as a carbon sink. We attempted to do this by comparing forest plots that were exposed to enhanced CO~2~ (eCO~2~) to those that were treated with ambient CO~2~ (aCO~2~). We found that enhanced CO~2~ tends to increase the rates of photosynthesis (Figure 1). Looking at the yearly photosynthesis averages for the years that we have data for, we were able to see that the average photosynthesis rates were higher in the plots that were treated with eCO~2~ compared to those treated with aCO~2~ (Figure 2). When we looked at the ratio of photosynthesis in eCO~2~ plots compared to the rates in aCO~2~ plots we found that for the three years that we have data, the ratio remained above one, indicating that the eCO~2~ plots had higher rates of photosynthesis than the aCO~2~ plots (Figure 1b). The increased rates of photosynthesis has some yearly fluctuations, but overall appears to be sustained over the 3 year period that we're looking at (Figure 1). This increase in photosynthesis at eCO~2~ levels allows us to hypothesize that as CO~2~ levels increase with global warming, this will most likely also increase the rates of photosynthesis that we see in plants around the world. However, more data that show how these rates might change in future years would be helpful in determining if the increased rates of photosynthesis with eCO~2~ are actually sustained (Figure 1). 
\
We also found that over the course of these three years, annual productivity, measured by litter fall, tended to increase for both the plots treated with aCO~2~ and eCO~2~ (Figure 3). Leaves were the largest component of this, followed by twigs, bark, and then seeds (Figure 3). This might lead us to believe that the trees in this plot are not yet mature and as a result of this they may still be growing and adding to their mass on a yearly basis. Along with this increase in productivity we also found that the plots treated with eCO~2~ tended to have less annual productivity than those treated with aCO~2~, however the amounts of productivity for these treatments was extremely close in 2023 (Figure 3). We found similar results when considering above ground net primary production (ANPP), where we found that over the course of the three years ANPP increased, and ANPP was greater in the plots with aCO~2~ than those with eCO~2~ (Figure 4). Based on the increased rates of photosynthesis in the plants treated with eCO~2~ we might have expected that this would lead to an increase in annual productivity as well as ANPP. However, this increase in photosynthesis without a corresponding increase in production could be taken to mean that the trees are storing more carbon for longer periods of time when there are elevated levels of CO~2~. 
\
Using this data we were also able to hypothesize the timing of a few defoliation events that occurred in this time period as a result of large insect outbreaks. When there are large insect outbreaks I would expect the insects to eat a large portion of the leaves off of the plants in this area. This would mean that there were then less leaves on the trees, and following these events there would likely be less leaves collected in the litter fall. For this reason, I believe that we would see evidence of these defoliation events in the dips in a graph showing the amounts of leaf litter that were collected at different points (Figure 5). There were three major defoliation events that I identified in this time period with one occurring in each year (Figure 5). There does not appear to be a huge difference in the defoliation events between the plots that were treated with aCO~2~ and those that were treated with eCO~2~ (Figure 5). The only potential difference that I see is that after the 2022 event and the 2023 event, it appears that the plots that were treated with aCO~2~ increased their production of leaves at a much higher rather than those treated with eCO~2~ (Figure 5). This would make sense when we consider the fact that annual productivity and ANPP were higher in the plots treated with aCO~2~ which could explain why the leaves on these trees grew back more quickly following these defoliation events.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\pagebreak

```{r photo, echo=FALSE, message=FALSE, warning=FALSE, fig.height=6.5, fig.width=6}

photo <- read.csv("data/face_photosynthesis.csv")

photo$Date <- paste(photo$Date, "01", sep= "_")
photo$Date2 <- as.Date(photo$Date, format="%b_%Y_%d")

library(doBy)
library(sciplot)

photo_co2 <- summaryBy(Anet ~ Date2 + treatment, data = photo,
                       FUN = c(mean, se), na.rm = TRUE)

cols <- c("pink", "cornflowerblue")

split.trt <- split(photo_co2, photo_co2$treatment)

photolab <- expression(Foliage~Photosynthesis~~(umols~m^-2~s^-1))

aco2.sub <- subset(photo_co2, treatment == "aCO2", select = c(Date2, treatment, Anet.mean))
eco2.sub <- subset(photo_co2, treatment == "eCO2", select = c(Date2, treatment, Anet.mean))

names(aco2.sub)[3] <- "aco2.mean"
names(eco2.sub)[3] <- "eco2.mean"

merge.data <- merge(aco2.sub, eco2.sub, by = "Date2")

merge.data$ratio <- merge.data$eco2.mean/merge.data$aco2.mean


ratiolab <- expression(Response~Ratio~~(eCO[2]/aCO[2]))

par(mfrow = c(2,1))
par(mar=c(2, 5, 1, 1))
with(split.trt$aCO2, plot(Date2, Anet.mean, type = 'l', col = "pink", ylim = c(14,30), 
                          xlab = "", ylab = photolab, xaxt = 'n')) 
with(split.trt$eCO2, lines(Date2, Anet.mean, type = 'l', col = "cornflowerblue")) 
legend("topright", legend = c("aCO2","eCO2"), col = cols, lty = 1,
       cex = 1, bty = "n")
legend("bottomleft", legend = "A", bty = 'n')
par(mar=c(5, 5, 0, 1))
with(merge.data, plot(Date2, ratio, ylim = c(0.5, 1.5), type = 'l', ylab = ratiolab, 
                      xlab = "Year"))
abline(h=1, lty = 2)
legend("bottomleft", legend = "B", bty = 'n')

```
**Figure 1.** Response of the foliage photosynthesis to aCO~2~ and eCO~2~. Graph A examines the rates of foliage photosynthesis with aCO~2~ depicted by the pink line and eCO~2~ depicted by the blue line. Graph B shows the response ratio (indicated by eCO~2~ levels divided by aCO~2~ levels). The dotted line at y = 1 on this graph indicates where we would expect this ratio to fall if the trees responded to eCO~2~ and aCO~2~ at the same rates.


```{r photo mean, echo=FALSE, message=FALSE, warning=FALSE}

photo_co2$year <- format(photo_co2$Date2, format = "%Y")

year.mean <- with(photo_co2, tapply(Anet.mean, list(treatment, year), mean))

photo$year <- format(photo$Date2, format = "%Y")
photo2 <- summaryBy(Anet ~ treatment + year, data = photo,
                    FUN = c(mean, se), na.rm = TRUE)

photolab2 <- expression(Average~Photosynthesis~~(umols~m^-2~s^-1))

library(plotrix)
par(mfrow = c(1,1))
par(mar=c(5, 5, 1, 1))
b <- barplot(year.mean, beside = TRUE, col=cols[photo_co2$treatment], width = 0.5,
             ylim = c(0,37),ylab = photolab2, xlab = "Year")
legend("topright", legend = levels(photo_co2$treatment), pch = 16, col = cols, bty = "n")
plotCI(b, year.mean, uiw = photo2$Anet.se, add = TRUE, pch = NA)
abline(h=0)

```
**Figure 2.** Average photosynthesis rates for each year that data was collected for. The pink bars indicate the values for aCO~2~ and the blue bars indicate the values for eCO~2~. The lines on the graphs indicate the standard error for each of the averages.


```{r litter, echo=FALSE, message=FALSE, warning=FALSE}

lit_annual <- read.csv("data/lit_annual.csv")

library(ggplot2)

litter_co2 <- summaryBy(twig+bark+seed+leaf ~ year+treatment, data = lit_annual, FUN =mean,
                        keep.names = TRUE)

litter_fum <- droplevels(litter_co2[litter_co2$year != 2020,])

aco2.twig <- subset(litter_fum, treatment == "aco2", select = c(year, treatment, twig))
aco2.twig$Component <- c("twig")
colnames(aco2.twig)[3] <- "amount"
aco2.bark <- subset(litter_fum, treatment == "aco2", select = c(year, treatment, bark))
aco2.bark$Component <- c("bark")
colnames(aco2.bark)[3] <- "amount"
aco2.seed <- subset(litter_fum, treatment == "aco2", select = c(year, treatment, seed))
aco2.seed$Component <- c("seed")
colnames(aco2.seed)[3] <- "amount"
aco2.leaf <- subset(litter_fum, treatment == "aco2", select = c(year, treatment, leaf))
aco2.leaf$Component <- c("leaf")
colnames(aco2.leaf)[3] <- "amount"
eco2.twig <- subset(litter_fum, treatment == "eco2", select = c(year, treatment, twig))
eco2.twig$Component <- c("twig")
colnames(eco2.twig)[3] <- "amount"
eco2.bark <- subset(litter_fum, treatment == "eco2", select = c(year, treatment, bark))
eco2.bark$Component <- c("bark")
colnames(eco2.bark)[3] <- "amount"
eco2.seed <- subset(litter_fum, treatment == "eco2", select = c(year, treatment, seed))
eco2.seed$Component <- c("seed")
colnames(eco2.seed)[3] <- "amount"
eco2.leaf <- subset(litter_fum, treatment == "eco2", select = c(year, treatment, leaf))
eco2.leaf$Component <- c("leaf")
colnames(eco2.leaf)[3] <- "amount"

stack.data <- rbind(aco2.twig, aco2.bark, aco2.seed, aco2.leaf, eco2.twig, eco2.bark,
                    eco2.seed, eco2.leaf)

litlab <- expression(Annual~Productivity~~(g~C~m^-2~year^-1))

ggplot(data=stack.data, aes(x=treatment, y=amount, fill=Component)) + 
  geom_bar(stat="identity") +
  facet_grid(~year) +
  ylab(litlab) +
  xlab("Treatment") +
  labs(color = "Component")

```
**Figure 3.** Annual productivity of trees treated with either aCO~2~ or eCO~2~. Different tree components are depicted by the different colors in the stacked bar graph with coral indicating bark, green indicating leaves, blue indicating seeds, and purple indicating twigs. The bars are broken up between the three years that data was collected for as well as the two different treatments, aCO~2~ or eCO~2~.


```{r anpp, echo=FALSE, message=FALSE, warning=FALSE}

wood_npp <- read.csv("data/wood_npp.csv")
lit_annual <- read.csv("data/lit_annual.csv")

colnames(wood_npp)[2]<-"ring"

wood_npp$treatment <- ifelse(wood_npp$treatment == "eCO2", "eco2",
                           "aco2")

litter_fum <- droplevels(lit_annual[lit_annual$year != 2020,])

combdata <- merge(litter_fum, wood_npp, by = c("year","ring","treatment"))

combdata$total <- (combdata$twig+ combdata$bark+ combdata$seed+ combdata$leaf+ 
                      combdata$wood_yr)


anpp <- summaryBy(total ~ year + treatment, data = combdata,
                       FUN = c(mean, se), na.rm = TRUE)


cols <- c("pink", "cornflowerblue")

year.mean2 <- with(anpp, tapply(total.mean, list(treatment, year), mean))

library(plotrix)
anpplab <- expression(ANPP~~(g~C~m^-2~year^-1))

par(mar=c(5, 5, 2, 2))
c <- barplot(year.mean2, beside = TRUE, col=cols[anpp$treatment], width = 0.5,
             ylim = c(0, 450), ylab = anpplab, xlab = "Year")
legend("topright", legend = levels(photo_co2$treatment), pch = 16, col = cols, bty = "n")
plotCI(c, year.mean2, uiw = anpp$total.se, add = TRUE, pch = NA)
abline(h=0)

```
**Figure 4.** ANPP of trees treated with aCO~2~ compared to those treated with eCO~2~ for the three years that we have data for. The pink bars indicate the values for aCO~2~ and the blue bars indicate the values for eCO~2~. The lines on the graphs indicate the standard error for each of the averages.


```{r bugs, echo=FALSE, message=FALSE, warning=FALSE}

litter <- read.csv("data/face_litter.csv")

library(doBy)
library(lubridate)

basket <- .2 
c_frac <- .47 

litter$Date <- as.Date(litter$Date, format = "%d/%m/%Y")
litter$year <- as.factor(year(litter$Date))
litter$treatment <- as.factor(ifelse(litter$ring == 2 | 
                                       litter$ring == 3 |
                                       litter$ring == 6, 
                                     "aco2", "eco2"))

line.num <- which.max(litter$twig)
litter <- litter[-line.num, ]

line.num <- which.max(litter$twig)
litter <- litter[-line.num, ]

litter_carbon <- data.frame(litter[,c(1:3, 8:9)], (litter[,4:7] * c_frac)/basket)

litter_mean_trap <- summaryBy(twig + bark+ seed + leaf ~ year + Date + ring +treatment,
                              data=litter_carbon, FUN=mean, na.rm=TRUE, keep.names = TRUE)

litter_mean_trap$Date3 <- as.numeric(litter_mean_trap$Date)
litter_bug <- summaryBy(twig+bark+seed+leaf ~ Date+treatment+Date3+year, data = litter_mean_trap,
                        FUN =mean,
                        keep.names = TRUE)

split.trt2 <- split(litter_bug, litter_bug$treatment)

cols <- c("pink", "cornflowerblue")
boxcol <- alpha("darkseagreen", alpha = .25)

ylabel <- expression(Leaf~Productivity~~(g~C~m^-2~year^-1))

par(mar=c(5, 5, 1, 1))
with(split.trt2$aco2, plot(Date3, leaf, type = 'l', col = "pink", 
                          xlab = "Year", ylab = ylabel, ylim = c(0,70), 
                          xaxt = 'n')) 
with(split.trt2$eco2, lines(Date3, leaf, type = 'l',
                            col = "cornflowerblue")) 
legend("topleft", legend = c("aCO2","eCO2"), col = cols, lty = 1,
       cex = 1, bty = "n")
axis(1, at=c(18635,19006,19370,19708), labels = c("2021", "2022", "2023","2024"))
rect(xleft = 18700, ybottom = 1, xright = 18850, ytop = 10, col = boxcol)
rect(xleft = 19050, ybottom = 0, xright = 19255, ytop = 10, col = boxcol)
rect(xleft = 19425, ybottom = 3.5, xright = 19555, ytop = 12, col = boxcol)

```
**Figure 5.** Leaf productivity of trees treated with aCO~2~ and eCO~2~ over the course of about 3 years. Leaf productivity of trees treated with aCO~2~ is depicted by the pink line and those treated with eCO~2~ are depicted by the blue line. The green boxes indicate places where I believe defoliation events to have occurred.

