---
title: FACE Project data analysis of photosynthesis and litterfall impacts of elevated
  CO2 rings
author: "MaryannWebb"
date: "4/16/2018"
output:
 word_document:
   reference_docx: manuscriptstyle.docx
---

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

# Photosynthesis
## Comparing Elevated CO^2 with Ambient CO^2
From 2021 to 2024, there was a constant trend of net photosynthesis being higher in CO2 FACE experiment than those with ambient CO2 (Figure 1). The overall trend remains constant with fluctutions of net photosynthesis between the years. Between 2022 and 2023, there are the largest gaps between photosynthesis levels and treatment. Despite annual fluctiontions, excess CO2 correlates to higher net photsyntheis values than those of ambient CO2. 

## Response Ratios
Response ratios are calculated by dividing elevated CO2 photosynthetic values by ambient CO2 values in order to see if carbon dioxide fertilization is sustained over time. The response ratio for this particular FACE experiment showed a varaition in fertilization between 2021 and 2024 (Figure 2). There are dramatic peaks and drops in response ratio which could indicate seasonal changes in leaf photosynthetic production if differences in the incident angle between seasons is very large, or if the FACE species is deciduous, thus only has leaves to photosynthesize in the summer months. 

#Annual Productivity 
##Ambient Carbon Dioxide Litterfall 
Annual Productivity in ambient treatments saw the greatest amount of carbon in 2021, then 2023, and the least amount in 2022 (Figure 3). Leaves make up the bulk of carbon biomass in litterfall, although it is close to twig biomass in 2021. Seeds make up the smallest percent of annual litterfall in all years. 

## Excess Carbon Dioxide Litterfall
Annual productivity measured by litterfall saw the greatest biomass in 2022 in CO2 treated rings (Figure 4). Similar to what is seen in ambient litterfall (Figure 3), leaves make up the majority of the biomass collected. Twigs are second to leaves in the percent of biomass collected, and sees an increase in amount in 2022. 

##Comparing Ambient and Excess Carbon Dioxide Litterfall 
Ambient treatments saw small decreases in the amount of carbon (from litterfall) from 2021 to 2023, while excess treatments saw an increase. In 2021, the first year of the fully functional FACE experiment, ambient treatments has more carbon in litterfall than the excess treatments (Figure 5). It is likely that it took more than one year of the carbon treatment for the trees to show a measurable impact on carbon storage in the litterfall. 

#Above Ground Net Primary Production (ANPP)
## Comparing Ambient and Excecss Carbon Treatments 
Wood biomass was averaged by year, with the net gain in wood mass taken as the increase in mass between measurements. Adding this value to the litterfall values, we were able to calculate total ANPP for the two treatments. When treatment was in its first year of full function, there was a large difference in biomass with ambient being much higher than in excess carbon dioxide sites (Figure 6). In 2022 ambient sites lost biomass while excess sites surpassed biomass amounts seen in ambient sites in the same year. The final year saw the two treatments very close together in total amount of carbon, with ambient having slightly more biomass than excess carbon sites. These fluctuations indicate that although introducing carbon is having an effect on growth, the fluctuations in ambient growth indicate that there may be other environmental factors or disturbances that are not giving a clear picture of how the system is operating. 

#Defoliation Events
Tracking the amount of leaf litter collected over time shows if there are prominent defoliation events to consider at the sites. Defoliation was equal between the two treatments in 2020 and 2021, but the amount of leaf litter increased dramatically between the two years, showing a possible defoliation event. In 2022 there was a larger increas in leaf litter in ambient treatments and a small increase in litter in excess treatments. The two treatments reach similar levels of leaf loss in 2023, with ambient treatments still seeing more litter (Figure 7). This is the opposite of what one would expect as excess carbon in theory instigates growth in trees, thus increasing leaf production. It is possible that more leaves were actually digested in the excess rings as leaves would be more nutritous with excess carbon, creating an imbalance in the amount of leaves collected between the two plots. 

```{r photosynthesis, echo=FALSE, warning=FALSE, message=FALSE}
photo<- read.csv("face_photosynthesis.csv")

library(doBy)
library(sciplot)
library(plotrix)
mycol=c("navy", "red")
photo$Date <-paste(photo$Date, "01", sep="_")
photo$date<-as.Date(photo$Date, format="%b_%Y_%d")
photo_co2<-summaryBy(Anet ~ date + treatment, data=photo, FUN=c(mean, se), na.rm=TRUE)
ambient <- data.frame(photo_co2[c(1,3,5,7,9,11,13,15,17,19),])
excess <- data.frame(photo_co2[c(2,4,6,8,10,12,14,16,18,20),])
par(mfrow=c(2,1), xpd=TRUE, mar=c(0,0,0,0), oma=c(5,5,1,1))
plot(Anet.mean ~ date, data=ambient,
     col="navy", pch=18, xaxt='n', ylim=c(15,30))
with(ambient, arrows(date, Anet.mean, date, Anet.mean+Anet.se, angle=90, length=0.03, col="navy"))
with(ambient, arrows(date, Anet.mean, date, Anet.mean-Anet.se, angle=90, length=0.03, col="navy"))
plot(Anet.mean ~ date, data=excess,
     col="red", pch=18, ylim=c(15,30), xlab="Year")
with(excess, arrows(date, Anet.mean, date, Anet.mean+Anet.se, angle=90, length=0.03, col="red"))
with(excess, arrows(date, Anet.mean, date, Anet.mean-Anet.se, angle=90, length=0.03, col="red"))
```
Figure 1: Net photosynthesis measured from both ambient and excess carbon dioxide treatments from 2021 to 2024. 

```{r response ratio, message=FALSE, warning=FALSE, echo=FALSE}
merge(ambient, excess, by="date")
joint<- data.frame(merge(ambient, excess, by="date"))
colnames(joint)[3]<-"Anet.mean.aco2"
colnames(joint)[6]<-"Anet.mean.eco2"

joint$ratio <- joint$Anet.mean.eco2/joint$Anet.mean.aco2
#does response ratio have se bars if not a mean?

plot(ratio ~ date, data=joint, type='l',
     pch=18, xlab="Year", ylab="Response Ratio", col="navy", ylim=c(1.00, 1.40))


```
Figure 2: Response ratio (carbon dioxide/ambient) net photosynthetic values from 2021 to 2024. 
\
```{r ambient_litter, warning=FALSE, echo=FALSE, message=FALSE}
litter_annual<- read.csv("litter_annual.csv")
library(doBy)
library(lubridate)
library(magicaxis)
library(plotrix)
litter_colors<-c("red", "yellow", "forest green", "dodger blue")
all_final <- droplevels(litter_annual[litter_annual$year !=2020,])

all_final$total <- all_final$twig+all_final$bark+all_final$leaf+all_final$seed

all_final$treatment <- as.factor(ifelse(all_final$ring ==2 |
                                       all_final$ring==3 |
                                       all_final$ring==6, 
                                     "aCO2", "eCO2"))
#aCO2

aCO2 <- droplevels(all_final[all_final$treatment !="eCO2",])

aCO2_annual<- summaryBy(twig + bark + leaf +seed ~ year, data=aCO2, FUN=c(mean, sum))

aCO2_annual$total<- aCO2_annual$twig.mean+aCO2_annual$bark.mean+aCO2_annual$leaf.mean+aCO2_annual$seed.mean
write.csv(aCO2_annual, "aCO2_annual.csv")

aCO2_data <- read.table(text="2021, 2022, 2023
                        bark 269.3872 322.6993 308.0129
                        twigs 953.0363 402.6206 407.7248
                        leaves 1014.685 1445.589 1511.598
                        seeds 174.09732 28.83998 17.97695")
aco2_barplot<- barplot(as.matrix(aCO2_data), col=litter_colors,
        xaxt='n', xlab="Year", ylab="Percent Carbon")
legend("topright", legend=c("Bark", "Twigs", "Leaves", "Seeds"), col=litter_colors, pch=20)
axis(1, at=aco2_barplot, labels=c("2021", "2022", "2023"))



```
Figure 3: Biomass in percent carbon by litterfall type collected in ambient carbon dioxide treatments from 2021 to 2023.

```{r eco2_photo, message=FALSE, warning=FALSE, echo=FALSE}
#eCO2

eCO2<- droplevels(all_final[all_final$treatment !="aCO2",])
eCO2_annual<- summaryBy(twig+bark+leaf+seed ~ year, data=eCO2, FUN=c(mean, sum))
eCO2_annual$total<- eCO2_annual$twig.mean+eCO2_annual$bark.mean+eCO2_annual$leaf.mean+eCO2_annual$seed.mean
write.csv(eCO2_annual, "eCO2_annual.csv")

eCO2_data<- read.table(text=" 2021, 2022, 2023
                       bark 295.6759 363.9051 407.7873
                       twigs 502.8030 825.1382 389.3327
                       leaves 951.1192 1056.7238 1425.1476
                       seed 215.07271 21.16285 12.89899")

barplot<-barplot(as.matrix(eCO2_data), col=litter_colors,
         xaxt='n', ylab="Percent Carbon", xlab="Year")
axis(1, at=barplot, labels=c("2021", "2022", "2023"))
legend("topright", legend=c("Bark", "Twigs", "Leaves", "seeds"), col=litter_colors, pch=20)

```
Figure 4: Biomass in percent carbon by litterfall type collected in excess carbon dioxide treatments from 2021 to 2023.
```{r both treatments litterfall, message=FALSE, warning=FALSE, echo=FALSE}
litter_colors<-c("red", "yellow", "forest green", "dodger blue")
total_data_trt<- read.table(text="2021, 2022, 2023
                        aCO2 2411.206 2199.748 2245.312
                        eCO2 1964.671 2266.930 2235.167")
aco2_barplot<-barplot(as.matrix(total_data_trt), col=litter_colors, xaxt='n', xlab="Year", ylab="Total Carbon")
legend("topright", legend=c("Ambient CO2", "Excess CO2"), col=litter_colors, pch=20)
```
Figure 5: Biomass in percent carbon collected as litterfall (includeing leaves, twigs, seeds, and bark) in both ambient and excess carbon treatments from 2021 to 2023

```{r ANPP, echo=FALSE, message=FALSE, warning=FALSE}
wood<-read.csv("wood_npp.csv")
litter_annual<- read.csv("litter_annual.csv")
aCO2_annual<-read.csv("aCO2_annual.csv")
eCO2_annual<- read.csv("eCO2_annual.csv")
library(plotrix)
library(magicaxis)
library(doBy)
wood_aco2<-droplevels(wood[wood$treatment !="eCO2",c(1,5)])
wood_aco2_annual<-summaryBy(wood_yr ~ year, data=wood_aco2, FUN=mean)
aco2_wood_litter<-merge(wood_aco2_annual, aCO2_annual[,2:6], by="year")
aco2_wood_litter$total_biomass_a<-(aco2_wood_litter$wood_yr.mean+aco2_wood_litter$twig.mean+
                                   aco2_wood_litter$bark.mean+aco2_wood_litter$leaf.mean+
                                   aco2_wood_litter$seed.mean)
wood_eco2<-droplevels(wood[wood$treatment !="aCO2", c(1,5)])
wood_eco2_annual<-summaryBy(wood_yr ~ year, data=wood_eco2, FUN=mean)
eco2_wood_litter<-merge(wood_eco2_annual, eCO2_annual[,2:6], by="year")
eco2_wood_litter$total_biomass_e<-(eco2_wood_litter$wood_yr.mean + eco2_wood_litter$twig.mean
                                 +eco2_wood_litter$bark.mean+eco2_wood_litter$leaf.mean
                                 +eco2_wood_litter$seed.mean)

plot(total_biomass_a~year, data=aco2_wood_litter, col="red", ylim=c(1900,2500),
     xaxt='n', ylab="Total Carbon (Biomass)", xlab="Year", pch=19)
points(total_biomass_e~year, data=eco2_wood_litter, col="navy", pch=19)
magaxis(side=1, majorn=3, minorn=0, labels=TRUE)
legend("topright", legend=c("Excess", "Ambient"), pch=19, col=c("navy", "red"))

```

Figure 6:Total carbon, including litterfall and wood growth representing ANPP changes from 2021 to 2023 in both ambient and excess carbon treatments.

```{r bugs, echo=FALSE, message=FALSE, warning=FALSE}
litter_annual<-read.csv("litter_annual.csv")
library(doBy)
library(plotrix)
library(magicaxis)
leaf_colors<-c("red", "forest green")
litter_annual$treatment<- as.factor(ifelse(litter_annual$ring ==2 |
                                              litter_annual$ring==3 |
                                              litter_annual$ring==6, 
                                            "aCO2", "eCO2"))
leaf_litter<-droplevels(litter_annual[,c(1,6,7)])
leaf_litter_annual<-summaryBy(leaf~year+treatment, data=leaf_litter, FUN=mean)
plot(leaf.mean~year, data=leaf_litter_annual,
     col=leaf_colors[treatment], pch=19, ylab="Mean Leaf Litter Mass", xlab="Year", 
     xaxt='n')
magaxis(side=1, majorn=3, minorn=0, labels=TRUE)
legend("bottomright", legend=c("Ambient", "Excess"), pch=19, col=leaf_colors)
```
Figure 7: Defoliation events characterized by increased leaf litterfall biomass for both ambient and excess carbon treatments.
